---
title: "DspikeIn with TSE"
author: "Mitra Ghotbi"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
    highlight: tango
    df_print: tibble
    css: vignette-style.css  
fontsize: 11pt
mainfont: "Arial"
geometry: margin=1in
vignette: >
  %\VignetteIndexEntry{DspikeIn with TSE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---



```{r setup, include = FALSE}

knitr::opts_chunk$set(
  fig.path = "figures/",        
  dev = "png",                  
  dpi = 300,                     
  fig.width = 7,               
  fig.height = 4,               
  fig.align = "center",          
  out.width = "70%",           
  message = FALSE,              
  warning = FALSE               
)


```


#                Required Packages


```{r, eval=FALSE}

#                         INSTALL CRAN PACKAGES

# Install missing CRAN packages
install.packages(setdiff(
  c(
    "stats", "dplyr", "ggplot2", "flextable", "ggpubr",
    "randomForest", "ggridges", "ggalluvial", "tibble",
    "matrixStats", "RColorBrewer", "ape", "rlang",
    "scales", "magrittr", "phangorn", "igraph", "tidyr",
    "xml2", "data.table", "reshape2", "vegan", "patchwork", "officer"
  ),
  installed.packages()[, "Package"]
))

# Load CRAN packages
lapply(c(
  "stats", "dplyr", "ggplot2", "flextable", "ggpubr", "randomForest",
  "ggridges", "ggalluvial", "tibble", "matrixStats", "RColorBrewer",
  "ape", "rlang", "scales", "magrittr", "phangorn", "igraph", "tidyr",
  "xml2", "data.table", "reshape2", "vegan", "patchwork", "officer"
), library, character.only = TRUE)


#                         INSTALL BIOCONDUCTOR PACKAGES


# Install BiocManager if not installed
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")

# Install missing Bioconductor packages
BiocManager::install(setdiff(
  c(
    "phyloseq", "msa", "DESeq2", "ggtree", "edgeR",
    "Biostrings", "DECIPHER", "microbiome", "limma",
    "S4Vectors", "SummarizedExperiment", "TreeSummarizedExperiment"
  ),
  installed.packages()[, "Package"]
))

# Load Bioconductor packages
lapply(
  c(
    "phyloseq", "msa", "DESeq2", "edgeR", "Biostrings", "ggtree", "DECIPHER",
    "microbiome", "limma", "S4Vectors", "SummarizedExperiment", "TreeSummarizedExperiment"
  ),
  library,
  character.only = TRUE
)


#                         INSTALL DspikeIn FROM GITHUB


# To access the DspikeIn vignette for a detailed tutorial, use vignette("DspikeIn"), or browse all available vignettes with browseVignettes("DspikeIn").
devtools::install_github("mghotbi/DspikeIn", build_vignettes = TRUE, dependencies = TRUE)
library(DspikeIn)
browseVignettes("DspikeIn")
vignette("DspikeIn")

## or

if (!requireNamespace("devtools", quietly = TRUE)) install.packages("devtools")
devtools::install_github("mghotbi/DspikeIn")

# Load DspikeIn only if installed
if ("DspikeIn" %in% installed.packages()[, "Package"]) {
  library(DspikeIn)
} else {
  stop("DspikeIn installation failed. Check errors above.")
}
```


#    Acknowledgments
The development of the DspikeIn package was made possible through the generous and pioneering efforts of the R and Bioconductor communities. We gratefully acknowledge the developers and maintainers of the following open-source packages, whose tools and infrastructure underpin our work:
**Core infrastructure & data manipulation:**
methods, stats, utils, graphics, grDevices, data.table, dplyr, tibble, tidyr, reshape2, matrixStats, rlang, S4Vectors, grid, officer, xml2
**Statistical analysis & modeling:**
DESeq2, edgeR, limma, randomForest, microbiome
**Phylogenetics & microbiome structure:**
phyloseq, TreeSummarizedExperiment, SummarizedExperiment, phangorn, ape, DECIPHER, msa, Biostrings
**Network and graph analysis:**
igraph, ggraph
**Visualization & layout design:**
ggplot2, ggrepel, ggpubr, ggnewscale, ggalluvial, ggtree, ggtreeExtra, ggstar, ggridges, patchwork, scales, RColorBrewer, flextable

These tools collectively empowered us to build a reproducible, modular, and extensible platform for robust absolute abundance quantification in microbial community analysis.
We further acknowledge the broader scientific community working on absolute microbial quantification, spike-in calibration, and compositional data analysis, whose foundational insights directly informed the design and conceptual framework of DspikeIn.



#     DspikeIn
The DspikeIn package supports both phyloseq and TreeSummarizedExperiment formats to streamline microbial quantification across diverse experimental setups. It accommodates either a single spike-in taxon or synthetic community taxa with variable or equal spike-in volumes and copy numbers. The package offers a comprehensive suite of tools for AA quantification, addressing challenges through ten core functions: 1) validation of spiked species, 2) data preprocessing, 3) system-specific spiked species retrieval, 4) scaling factor calculation, 5) conversion to absolute abundance, 6) bias correction and normalization, 7) performance assessment, and 8) taxa exploration and filtering 9) network topology assessment 10) further analyses and visualization. 


#    DspikeIn requirements
DspikeIn works with 7 taxonomic ranks


```{r, eval=FALSE}

#             To remove strain from the taxonomic ranks

# DspikeIn works with 7 taxonomic ranks
# colnames(taxmat) <- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")

# Function to remove strain or bracketed info
remove_strain_info <- function(df) {
  pattern <- "Strain.*|strain.*|\\s*\\[.*?\\]|\\s*\\(.*?\\)"
  for (col in colnames(df)) {
    df[[col]] <- gsub(pattern, "", df[[col]])
    df[[col]] <- trimws(df[[col]])}
  return(df)}

# Apply to rowData of TSE
taxonomy <- as.data.frame(rowData(tse_16SOTU))

# Clean strain 
cleaned_taxonomy <- remove_strain_info(taxonomy)

#  drop "Strain" if present
if ("Strain" %in% colnames(cleaned_taxonomy)) {
  cleaned_taxonomy <- cleaned_taxonomy[, colnames(cleaned_taxonomy) != "Strain"]}

# Re-assign to TSE
rowData(tse_16SOTU) <- cleaned_taxonomy


#             To add species rank to the taxonomic ranks

# Make sure 'Genus' exists
if (!"Genus" %in% colnames(rowData(tse_16SOTU))) {
  stop("The 'Genus' column is missing in rowData.")}

# prepare taxonomy
taxonomy <- as.data.frame(rowData(tse_16SOTU))

# Handle missing genus labels
taxonomy$Genus[is.na(taxonomy$Genus) | taxonomy$Genus == ""] <- "Unknown"
taxonomy$Species <- paste0(taxonomy$Genus, "_OTU", seq_len(nrow(taxonomy)))

# Assign back to TSE
rowData(tse_16SOTU) <- taxonomy

```



#        Build TreeSummarizedExperiment file
for more information please refer to 
https://github.com/markrobinsonuzh/TreeSummarizedExperiment


```{r, eval=FALSE}


#                       Build TreeSummarizedExperiment (TSE)


otu <- read.csv("otu.csv", header = TRUE, sep = ",", row.names = 1)
otu_mat <- as.matrix(otu) # Convert to matrix
tax <- read.csv("tax.csv", header = TRUE, sep = ",", row.names = 1)
colnames(tax) <- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")
tax_mat <- as.matrix(tax) # Convert to matrix
meta <- read.csv("metadata.csv", header = TRUE, sep = ",", row.names = 1)
reference_seqs <- readDNAStringSet("dna-sequences.fasta", format = "fasta")
tse <- TreeSummarizedExperiment(
  assays = list(counts = otu_mat), # OTU table
  rowData = tax_mat, # Taxonomy information
  colData = meta, # Sample metadata
  rowTree = MyTree, # Phylogenetic tree
  rowSeqs = reference_seqs # Reference sequences)
```



Do all detected sample spike-in sequences cluster with the reference, and are their branch lengths statistically similar, supporting a common ancestor?

# spike-in validation 

All sample-derived sequences are forming a clade with the reference.
We look for a monophyletic grouping of spike-in OTUs
The clade is strongly supported (bootstrap around 100 percentage).
The branch lengths and distances are in a biologically plausible range.


```{r}
# Use the Neighbor-Joining method  based on a Jukes-Cantor distance matrix and plot the tree with bootstrap values.
# we compare the Sanger read of Tetragenococcus halophilus with the FASTA sequence of Tetragenococcus halophilus from our phyloseq object.

library(Biostrings)
library(TreeSummarizedExperiment)
library(SummarizedExperiment)
library(DspikeIn)

# Get path to external data folder
extdata_path <- system.file("extdata", package = "DspikeIn")
list.files(extdata_path)

data("physeq_16SOTU", package = "DspikeIn")
tse_16SOTU<-convert_phyloseq_to_tse(physeq_16SOTU)
tse_16SOTU <- tidy_phyloseq_tse(tse_16SOTU)

# Filter TSE object to keep only Bacteria and Archaea

tse_16SOTU <- tse_16SOTU[
  rowData(tse_16SOTU)$Kingdom %in% c("Bacteria", "Archaea"),]


library(Biostrings)
# 
# # Subset the TSE object to include only Tetragenococcus
# tetragenococcus_tse <- tse_16SOTU[
#   rowData(tse_16SOTU)$Genus == "Tetragenococcus" &
#     !is.na(rownames(tse_16SOTU)) &
#     rownames(tse_16SOTU) != "", ]
# 
# ref_sequences <- referenceSeq(tetragenococcus_tse)
# 
# # Convert to DNAStringSet if needed
# ref_sequences <- Biostrings::DNAStringSet(ref_sequences)
# Biostrings::writeXStringSet(ref_sequences, filepath = "Sample.fasta")




ref_fasta <- system.file("extdata", "Ref.fasta", package = "DspikeIn")
sample_fasta <- system.file("extdata", "Sample.fasta", package = "DspikeIn")



result <- validate_spikein_clade(
  reference_fasta = ref_fasta,
  sample_fasta = sample_fasta,
  bootstrap = 200,
  output_prefix = NULL)

# result$tree_plot
```


#     Did spike-ins behave as expected across all samples?
Tip labels=	OTU/ASV names
Branch length numbers=	Actual evolutionary distances (small = very similar)
Prevalence stars	How frequently the OTU occurs across samples
Blue bar ring=	Log10 mean abundance
Outer colored tiles=	The metadata variable you choose (e.g., Animal.type)


```{r}
#data("physeq_16SOTU", package = "DspikeIn")
tse_16SOTU
library(ggstar)
library(ggplot2)

# filter your object to only include spike-in taxa beforehand:
# change the OTU IDs for easy detection
# Big stars = detected in many samples
# Small stars = rarely detected
# log10(Mean Abundance) Bars= Color intensity reflects mean abundance.
# The log-transformed average abundance of each OTU across all samples where it appears.
# Extreme blue may signal unintended over-representation.
# Metadata Ring = factor of your interest e.g. Animal.type
# Each OTU is colored by where it was observed.
# Branch length numbers=	Actual evolutionary distances (small = very similar)


library(DspikeIn)
library(TreeSummarizedExperiment)

# ---- 1. Subset taxa where Genus is Tetragenococcus ----
spikein_tse <- tse_16SOTU[
  rowData(tse_16SOTU)$Genus == "Tetragenococcus", ]


# ---- 2. Diagnostic plot (tree-based) ----
ps <- plot_spikein_tree_diagnostic(
  obj = spikein_tse,
  metadata_var = "Animal.type",
  save_plot = FALSE )


```


#                Pre_processing
merges monophyletic ASVs/OTUs

#          Spiked species and related parameters for 16S 



```{r}

#                      PREREQUISITE FOR 16S & CALCULATE SPIKED %


# Define spiked species and related parameters**

library(flextable)
library(DspikeIn)
library(TreeSummarizedExperiment)
library(SummarizedExperiment)
library(dplyr)

# Define spike-in parameters
spiked_cells <- 1847
species_name <- spiked_species <- c("Tetragenococcus_halophilus", "Tetragenococcus_sp.")

merged_spiked_species <- "Tetragenococcus"

# If you prefer Genus-level matching
# species_name <- spiked_species <- c("Tetragenococcus")
# merged_spiked_species <- "Tetragenococcus"



# --------------------------------------------
# Subset taxa for the spike-in species
tse_16SOTU_spiked_taxa <- tse_16SOTU[rowData(tse_16SOTU)$Species %in% species_name, ]

# (OTU) IDs
hashcodes <- rownames(tse_16SOTU_spiked_taxa)

# --------------------------------------------
# Subset samples based on spike-in volume (e.g., "1" or "2")
tse_16SOTU_spiked_samples <- tse_16SOTU[, colData(tse_16SOTU)$spiked.volume %in% c("2", "1") ]

# --------------------------------------------
# Merge spiked OTUs using the DspikeIn function
# merges monophyletic ASVs/OTUs

# The function Pre_processing_species() merges ASVs/OTUs
# of a species using "sum" or "max" methods, preserving taxonomic,
# phylogenetic, and sequencing data.



library(DspikeIn)
library(TreeSummarizedExperiment)
library(SummarizedExperiment)
output_rds <- file.path(tempdir(), "merged_tse_sum.rds")

Spiked_16S_sum_scaled <- Pre_processing_species(
  tse_16SOTU_spiked_samples,
  species_name,
  merge_method = "sum",
  output_file = output_rds)

# --------------------------------------------
# Calculate spike-in recovery percentage
Perc <- calculate_spike_percentage(
  Spiked_16S_sum_scaled,
  merged_spiked_species,
  passed_range = c(0.1, 20))

```


#          Spiked species and related parameters for ITS


```{r, eval=FALSE}

#                      PREREQUISITE FOR ITS & CALCULATE SPIKED %


# Define spiked species and related parameters**


# Define the spiked species
# spiked_cells <- 733
# species_name <- spiked_species <- merged_spiked_species <- "Dekkera_bruxellensis"

# Subset taxa for spiked species
# Dekkera <- phyloseq::subset_taxa(
# physeq_ITSOTU,
# Species %in% species_name)

# hashcodes <- row.names(phyloseq::tax_table(Dekkera))

# Subset samples based on spiked volume
# physeq_ITSOTU_spiked <- phyloseq::subset_samples(physeq_ITSOTU, spiked.volume %in% c("2", "1"))

# if TSE format
# tse_ITSOTU <- convert_phyloseq_to_tse(physeq_ITSOTU)
# physeq_ITSOTU_spiked <- tse_ITSOTU[, tse_ITSOTU$spiked.volume %in% c("2", "1")]
```




#         Calculating Scaling Factors 


```{r}

#                      CALCULATE SCALING FACTORS

result <- calculate_spikeIn_factors(
  Spiked_16S_sum_scaled,
  spiked_cells = spiked_cells,
  merged_spiked_species = species_name)

# View extracted outputs
result$spiked_species_reads # Merged spiked species name
result$total_reads # Total reads detected for the spike
scaling_factors <- result$scaling_factors
head(scaling_factors) # Vector of scaling factors per sample
```


#                Convert relative counts to absolute counts


```{r}


#             Convert relative counts to absolute counts


# absolute counts=relative counts×sample-specific scaling factor

# Convert to absolute counts
library(DspikeIn)
absolute <- convert_to_absolute_counts(Spiked_16S_sum_scaled, scaling_factors)

# Extract processed data
absolute_counts <- absolute$absolute_counts
tse_absolute <- absolute$obj_adj

tse_absolute <- tidy_phyloseq_tse(tse_absolute)

# View absolute count data
head(tse_absolute)
```


#     Summary Stat


```{r}


#                      CALCULATE SPIKE PERCENTAGE & summary stat

# ** Calculate spike percentage & Generate summary statistics for absolute counts**

# Generate summary statistics for absolute counts
post_eval_summary <- calculate_summary_stats_table(absolute_counts)

# You may want to Back normal to check calculation accuracy
# the scaling factor was computed based on spiked species reads and fixed cell count.
# Multiplying the spiked species read count by this scaling factor restores the exact spiked cell count.
# lets check it
# BackNormal <- calculate_spike_percentage(
#  tse_absolute,
#  merged_spiked_species,
#  passed_range = c(0.1, 20)
# )
```


#           spiked species retrieval is system-dependent
The goal is to identify the range where, for example, the evenness of your community remains independent of spiked species retrieval—meaning the p-value should not be significant, and the R² value should be low, indicating minimal influence.
Hill number is interpretable as "effective diversity" (number of abundant species).

```{r}
# The acceptable range of spiked species retrieval is system-dependent
# Spiked species become centroid of the community (Distance to Centroid)
# Spiked species become dominant and imbalance the community (Evenness)

# What range of spiked species retrieval is appropriate for your system?
# Calculate Pielou's Evenness using Shannon index and species richness (Observed)
# Hill number q = 1 = exp(Shannon index), representing the effective number of equally abundant species. Hill is weighted by relative abundance, so dominant species have more influence.
# Unlike Pielou's evenness, this metric is not normalized by richness and it shows Effective number of common species

library(TreeSummarizedExperiment)
library(SummarizedExperiment)
library(dplyr)
library(tibble)
library(microbiome)
library(mia)
library(vegan)
library(S4Vectors)
library(ggplot2)


# --- 1. Extract current metadata from TSE ---
metadata <- colData(tse_absolute) %>%
  as.data.frame() %>%
  rownames_to_column("Sample")

# --- 2. Add spike-in reads (Perc) ---
# Ensure Perc has 'Sample' column and matching format
metadata <- dplyr::left_join(metadata, Perc, by = "Sample")

# --- 3. Estimate alpha diversity indices and extract ---
tse_absolute <- mia::addAlpha(
  tse_absolute,
  index = c("observed", "shannon", "pielou", "hill")
)

alpha_df <- colData(tse_absolute) %>%
  as.data.frame() %>%
  rownames_to_column("Sample") %>%
  select(Sample, observed, shannon, pielou, hill)

metadata <- dplyr::left_join(metadata, alpha_df, by = "Sample")

# --- 4. Compute Bray-Curtis distance to centroid ---
otu_mat <- assay(tse_absolute, "counts")
otu_mat <- t(otu_mat)  # samples as rows
otu_mat_rel <- vegan::decostand(otu_mat, method = "total")

centroid_profile <- colMeans(otu_mat_rel)

dist_to_centroid <- apply(otu_mat_rel, 1, function(x) {
  vegan::vegdist(rbind(x, centroid_profile), method = "bray")[1]
})

# Match and assign distances
metadata$Dist_to_Centroid <- dist_to_centroid[metadata$Sample]

# --- 5. Assign updated metadata to TSE ---
metadata <- column_to_rownames(metadata, var = "Sample")
metadata <- metadata[colnames(tse_absolute), , drop = FALSE]  # ensure correct order and size

colData(tse_absolute) <- S4Vectors::DataFrame(metadata)


# 4. Regression Plots: Diversity vs. Spike-in Reads
# ============================================================================

# Pielou’s Evenness
plot_object_pielou <- regression_plot(
  data = metadata,
  x_var = "pielou",
  y_var = "Spiked_Reads",
  custom_range = c(0.1, 20, 30, 40, 50, 60, 100),
  plot_title = "Pielou’s Evenness vs. Spike-in Reads"
)

# Hill Number (q = 1)
plot_object_hill <- regression_plot(
  data = metadata,
  x_var = "hill",
  y_var = "Spiked_Reads",
  custom_range = c(0.1, 10, 20, 30, 100),
  plot_title = "Hill Number (q = 1) vs. Spike-in Reads"
)

# Distance to Centroid
plot_object_centroid <- regression_plot(
  data = metadata,
  x_var = "Dist_to_Centroid",
  y_var = "Spiked_Reads",
  custom_range = c(0.1, 20, 30, 40, 50, 60, 100),
  plot_title = "Distance to Global Centroid vs. Spike-in Reads"
)


# Interpretation
# ============================================================================
# - Pielou's evenness is normalized by richness; useful for detecting imbalance.
# - Hill number q = 1 gives effective number of common species; sensitive to dominance.
# - Distance to centroid in full Bray-Curtis space shows deviation from the average community.

```


#            Calculate the percentage of spiked species retrieval


```{r}
# * Calculate the percentage of spiked species retrieval per sample*

library(mia)         
library(dplyr)
library(SummarizedExperiment)

# Subset TSE to remove blanks 
tse_absolute_filtered <- tse_absolute[, colData(tse_absolute)$sample.or.blank != "blank"]

# Calculate spike-in retrieval percentage
result_perc <- calculate_spike_percentage(
  tse_absolute_filtered,
  merged_spiked_species,
  passed_range = c(0.1, 20)
)

# Generate conclusion report
conc <- conclusion(
  tse_absolute_filtered,
  merged_spiked_species,
  max_passed_range = 20,
  output_path = output_path)

conc$full_report

# Filter to keep only the samples that passed
passed_samples <- result_perc$Sample[result_perc$Result == "passed"]

# Subset the TSE object to include only passed samples
tse_passed <- tse_absolute_filtered[, colnames(tse_absolute_filtered) %in% passed_samples]
dim(tse_passed)

```



#  Abundance-based Core microbiome


```{r, out.width="70%"}

tse_absolute <- absolute$obj_adj
pps_Abs <- DspikeIn::get_long_format_data(tse_absolute)

# calculation for relative abundance needs sum of total reads
# total_reads <- sum(pps_Abs$Abundance)

# Generate an alluvial plot

alluvial_plot_abs <- alluvial_plot(
  data = pps_Abs,
  axes = c("Host.genus", "Ecoregion.III","Diet", "Animal.ecomode"),
  abundance_threshold = 10000,
  fill_variable = "Class",
  silent = TRUE,
  abundance_type = "absolute",
  top_taxa = 15,
  text_size = 4,
  legend_ncol = 1,
  custom_colors = DspikeIn::color_palette$light_MG # Use the color palette from DspikeIn
)


```



#                 Data transform & Normalization

you may select to transform your data befor moving forward with Differential Abundance


```{r}
# you may need to normalize/transform your data to reduce biases

ps <- tse_16SOTU

# TC Normalization
result_TC <- normalization_set(ps, method = "TC", groups = "Host.species")
normalized_ps_TC <- result_TC$dat.normed
scaling_factors_TC <- result_TC$scaling.factor

# UQ Normalization
result_UQ <- normalization_set(ps, method = "UQ", groups = "Host.species")
normalized_ps_UQ <- result_UQ$dat.normed
scaling_factors_UQ <- result_UQ$scaling.factor

# Median Normalization
result_med <- normalization_set(ps, method = "med", groups = "Host.species")
normalized_ps_med <- result_med$dat.normed
scaling_factors_med <- result_med$scaling.factor

# DESeq Normalization
ps_n <- remove_zero_negative_count_samples(ps)
result_DESeq <- normalization_set(ps_n, method = "DESeq", groups = "Animal.type")
normalized_ps_DESeq <- result_DESeq$dat.normed
scaling_factors_DESeq <- result_DESeq$scaling.factor

# Poisson Normalization
result_Poisson <- normalization_set(ps, method = "Poisson", groups = "Host.genus")
normalized_ps_Poisson <- result_Poisson$dat.normed
scaling_factors_Poisson <- result_Poisson$scaling.factor

# Quantile Normalization
result_QN <- normalization_set(ps, method = "QN")
normalized_ps_QN <- result_QN$dat.normed
scaling_factors_QN <- result_QN$scaling.factor

# TMM Normalization
result_TMM <- normalization_set(ps, method = "TMM", groups = "Animal.type")
normalized_ps_TMM <- result_TMM$dat.normed
scaling_factors_TMM <- result_TMM$scaling.factor

# CLR Normalization
result_clr <- normalization_set(ps, method = "clr")
normalized_ps_clr <- result_clr$dat.normed
scaling_factors_clr <- result_clr$scaling.factor

# Rarefying
result_rar <- normalization_set(ps, method = "rar")
normalized_ps_rar <- result_rar$dat.normed
scaling_factors_rar <- result_rar$scaling.factor

# CSS Normalization
result_css <- normalization_set(ps, method = "css")
normalized_ps_css <- result_css$dat.normed
scaling_factors_css <- result_css$scaling.factor

# TSS Normalization
result_tss <- normalization_set(ps, method = "tss")
normalized_ps_tss <- result_tss$dat.normed
scaling_factors_tss <- result_tss$scaling.factor

# RLE Normalization
result_rle <- normalization_set(ps, method = "rle")
normalized_ps_rle <- result_rle$dat.normed
scaling_factors_rle <- result_rle$scaling.factor

```




 Ridge plot


```{r, out.width="50%"}
rf_tse <- RandomForest_selected(
  tse_16SOTU,
  response_var = "Host.genus",
  na_vars = c("Habitat", "Ecoregion.III", "Host.genus", "Diet")
)

ridge_tse<- ridge_plot_it(rf_tse, taxrank = "Family", top_n = 10) + scale_fill_manual(values = DspikeIn::color_palette$cool_MG)
# ridge_physeq

result_css <- normalization_set(rf_tse, method = "css")
normalized_ps_css <- result_css$dat.normed

ridge_tse <- ridge_plot_it(normalized_ps_css, taxrank = "Family", top_n = 10) + scale_fill_manual(values = DspikeIn::color_palette$cool_MG)

# ridge_physeq
```



#                  Further analysis 

remove the spike-in sp before further analysis

```{r}
library(mia)
library(dplyr)
library(SummarizedExperiment)

# ---Remove unwanted taxa by Genus, Family, or Order ---
tse_filtered <- tse_absolute[
  rowData(tse_absolute)$Genus != "Tetragenococcus" &
  rowData(tse_absolute)$Family != "Chloroplast" &
  rowData(tse_absolute)$Order != "Chloroplast", ]

tse_caudate <- tse_filtered[, colData(tse_filtered)$Clade.Order == "Caudate"]
genera_keep <- c("Desmognathus", "Plethodon", "Eurycea")
tse_three_genera <- tse_caudate[, colData(tse_caudate)$Host.genus %in% genera_keep]
tse_blue_ridge <- tse_three_genera[, colData(tse_three_genera)$Ecoregion.III == "Blue Ridge"]
tse_desmog <- tse_blue_ridge[, colData(tse_blue_ridge)$Host.genus == "Desmognathus"]

# --- Differential Abundance with DESeq2 ---
results_DESeq2 <- perform_and_visualize_DA(
  obj = tse_desmog,
  method = "DESeq2",
  group_var = "Host.taxon",
  contrast = c("Desmognathus monticola", "Desmognathus imitator"),
  output_csv_path = "DA_DESeq2.csv",
  target_glom = "Genus",
  significance_level = 0.05)


head(results_DESeq2$results)
results_DESeq2$obj_significant
# Optional: Visualization
# results_DESeq2$plot
# results_DESeq2$bar_plot


```

# Relative abundance

```{r}
# Relative abundance
#  Extract OTU matrices from TSE Relative counts
# --------------------------------------------

# ---Remove unwanted taxa by Genus, Family, or Order ---
tse_filtered <- tse_16SOTU[
  rowData(tse_16SOTU)$Genus != "Tetragenococcus" &
  rowData(tse_16SOTU)$Family != "Chloroplast" &
  rowData(tse_16SOTU)$Order != "Chloroplast", ]

tse_caudate <- tse_filtered[, colData(tse_filtered)$Clade.Order == "Caudate"]
genera_keep <- c("Desmognathus", "Plethodon", "Eurycea")
tse_three_genera <- tse_caudate[, colData(tse_caudate)$Host.genus %in% genera_keep]
tse_blue_ridge <- tse_three_genera[, colData(tse_three_genera)$Ecoregion.III == "Blue Ridge"]
tse_desmog_rel <- tse_blue_ridge[, colData(tse_blue_ridge)$Host.genus == "Desmognathus"]

# --- Differential Abundance with DESeq2 ---
results_DESeq2_rel <- perform_and_visualize_DA(
  obj = tse_desmog_rel,
  method = "DESeq2",
  group_var = "Host.taxon",
  contrast = c("Desmognathus monticola", "Desmognathus imitator"),
  output_csv_path = "DA_DESeq2.csv",
  target_glom = "Genus",
  significance_level = 0.05)


head(results_DESeq2_rel$results)
results_DESeq2_rel$obj_significant
# Optional: Visualization
# results_DESeq2$plot
# results_DESeq2$bar_plot


```


# Turnover (Presence/absence analysis)
taxonomic detection consistency between RA and AA OTU tables
Presence/absence analysis to detect concordance between AA and RA profiles

```{r}
# --------------------------------------------
#  Extract OTU matrices from TSE objects
# --------------------------------------------

otu_rel <- assay(tse_desmog_rel, "counts") 
otu_abs <- assay(tse_desmog, "counts")        

# Make sure samples are rows and taxa are columns
otu_rel <- t(otu_rel)
otu_abs <- t(otu_abs)

# Ensure they are matrices
otu_rel <- as.matrix(otu_rel)
otu_abs <- as.matrix(otu_abs)

# --------------------------------------------
# 2. Convert to Presence/Absence 
# --------------------------------------------
otu_rel_pa <- (otu_rel > 0) * 1
otu_abs_pa <- (otu_abs > 0) * 1

# --------------------------------------------
# 3. Identify common samples & taxa
# --------------------------------------------
shared_samples <- intersect(rownames(otu_rel_pa), rownames(otu_abs_pa))
shared_taxa <- intersect(colnames(otu_rel_pa), colnames(otu_abs_pa))

# Subset to common set
otu_rel_pa <- otu_rel_pa[shared_samples, shared_taxa]
otu_abs_pa <- otu_abs_pa[shared_samples, shared_taxa]

# --------------------------------------------
# 4. Compare presence/absence profiles
# --------------------------------------------
shared_pa <- (otu_rel_pa + otu_abs_pa) == 2
total_present <- (otu_rel_pa + otu_abs_pa) >= 1

# Calculate percent agreement (global)
percent_shared <- sum(shared_pa) / sum(total_present)

cat("Percent of shared taxa detections (AA vs RA):",
    round(100 * percent_shared, 1), "%\n")


```


# Differential abundance (Single and Multilayer Pairwise)

```{r, eval=FALSE}


# DspikeIn: Differential Abundance Examples
# Using perform_and_visualize_DA() with multiple contrast scenarios

# 1. Single Contrast

res_single <- perform_and_visualize_DA(
  obj = tse_16SOTU,
  method = "DESeq2",
  group_var = "Diet",
  contrast = c("Insectivore", "Carnivore"),
  target_glom = "Genus")



# 2. Single Factor — All Pairwise Contrasts
col_df <- as.data.frame(colData(tse_16SOTU))
col_df$Host.taxon <- factor(make.names(as.character(col_df$Host.taxon)))
colData(tse_16SOTU)$Host.taxon <- col_df$Host.taxon

# Get unique DESeq2-compatible factor levels
host_levels <- levels(col_df$Host.taxon)
print(host_levels)

# contrast list
contrast_named <- list(
  Host.taxon = combn(host_levels, 2, simplify = FALSE))

# multiple pairwise contrasts
res_multi <- perform_and_visualize_DA(
  obj = tse_16SOTU,
  method = "DESeq2",
  group_var = "Host.taxon",
  contrast = contrast_named,
  target_glom = "Genus")



# 3. Single Factor — Selected Contrasts

contrast_list <- list(
  c("Insectivore", "Carnivore"),
  c("Omnivore", "Herbivore"))

res_selected <- perform_and_visualize_DA(
  obj = tse_16SOTU,
  method = "DESeq2",
  group_var = "Diet",
  contrast = contrast_list,
  global_fdr = TRUE)



# 4. Multiple Factors — Selected Contrasts

contrast_named <- list(
  Diet = list(
    c("Insectivore", "Carnivore"),
    c("Omnivore", "Carnivore") ),
  Animal.type = list(
    c("Frog", "Salamander") ))

res_multi_factor <- perform_and_visualize_DA(
  obj = tse_16SOTU,
  method = "DESeq2",
  contrast = contrast_named,
  target_glom = "Genus",
  significance_level = 0.01,
  global_fdr = TRUE)



# 5. Multiple Factors — all pairwise contrasts

colData(tse_16SOTU)$Host.taxon <- droplevels(factor(colData(tse_16SOTU)$Host.taxon))
colData(tse_16SOTU)$Habitat <- droplevels(factor(colData(tse_16SOTU)$Habitat))
host_levels <- levels(colData(tse_16SOTU)$Host.taxon)
habitat_levels <- levels(colData(tse_16SOTU)$Habitat)

# Define pairwise contrasts as a named list
contrast_named <- list(
  Host.taxon = combn(host_levels, 2, simplify = FALSE),
  Habitat = combn(habitat_levels, 2, simplify = FALSE))

# Define ComboGroup variable
colData(tse_16SOTU)$ComboGroup <- factor(interaction(
  colData(tse_16SOTU)$Animal.type,
  colData(tse_16SOTU)$Diet,
  drop = TRUE))



# --- Run multi-factor DESeq2 comparisons ---
results_multi_factor <- perform_and_visualize_DA(
  obj = tse_16SOTU,
  method = "DESeq2",
  contrast = contrast_named,
  target_glom = "Genus",
  significance_level = 0.01)

# --- Combined Group Interactions ---
colData(tse_16SOTU)$ComboGroup <- factor(interaction(
  colData(tse_16SOTU)$Animal.type,
  colData(tse_16SOTU)$Diet,
  drop = TRUE))

# --- Define custom interaction contrasts ---
contrast_list <- list(
  c("Salamander.Insectivore", "Lizard.Insectivore"),
  c("Salamander.Carnivore", "Snake.Carnivore"),
  c("Salamander.Carnivore", "Frog.Carnivore"))

# --- Run combined group comparisons ---
res_combo <- perform_and_visualize_DA(
  obj = tse_16SOTU,
  method = "DESeq2",
  group_var = "ComboGroup",
  contrast = contrast_list,
  target_glom = "Genus",
  global_fdr = TRUE)

```



#                     Visualization


```{r}

#   Visualization of community composition

# Subset rows where Genus is not Tetragenococcus and not NA
keep_taxa <- !is.na(rowData(tse_16SOTU)$Genus) & rowData(tse_16SOTU)$Genus != "Tetragenococcus"
# Subset the TSE by keeping only those taxa (rows)
tse_filtered <- tse_16SOTU[keep_taxa, ]
# Exclude blanks
tse_filtered <- tse_filtered[, colData(tse_filtered)$sample.or.blank != "blank"]
# subset Salamander samples
tse_sal <- tse_filtered[, colData(tse_filtered)$Animal.type == "Salamander"]
Prok_OTU_sal <- tidy_phyloseq_tse(tse_sal)

TB<-taxa_barplot(
  Prok_OTU_sal,
  target_glom = "Genus",
  custom_tax_names = NULL,
  normalize = TRUE,
  treatment_variable = "Habitat",
  abundance_type = "relative",
  x_angle = 25,
  fill_variable = "Family",
  facet_variable = "Diet",
  top_n_taxa = 20,
  palette = DspikeIn::color_palette$mix_MG,
  legend_size = 11,
  legend_columns = 1,
  x_scale = "free",
  xlab = NULL)

```



#                 Microbial dynamics & NetWork comparision


```{r}

# 1. Initialization and loading NetWorks for Comparision


# library(SpiecEasi)
# library(ggnet)
# library(igraph)
library(DspikeIn)
library(tidyr)
library(dplyr)
library(ggpubr)
library(igraph)

# To create a microbial co-occurrence network, you can refer to the SpiecEasi package available at:
# SpiecEasi GitHub Repository  https://github.com/zdk123/SpiecEasi

# herp.Bas.rel.f is a merged phyloseq object for both bacterial and fungal domains
# spiec.easi(herp.Bas.rel.f, method='mb', lambda.min.ratio=1e-3, nlambda=250,pulsar.select=TRUE )

Complete <- load_graphml("Complete.graphml")
NoBasid <- load_graphml("NoBasid.graphml")
NoHubs <- load_graphml("NoHubs.graphml")

# result <- weight_Network(graph_path = "Complete.graphml")
# result

result_kk <- degree_network(
  graph_path = load_graphml("Complete.graphml"),
  save_metrics = TRUE,
  layout_type = "stress"
)
# print(result_kk$plot)

```



#            Network Topological Metrics


```{r}

# 2. Metrics Calculation


result_Complete <- node_level_metrics(Complete)
result_NoHubs <- node_level_metrics(NoHubs)
result_NoBasid <- node_level_metrics(NoBasid)

Complete_metrics <- result_Complete$metrics
Nohub_metrics <- result_NoHubs$metrics
Nobasid_metrics <- result_NoBasid$metrics

Complete_metrics <- data.frame(lapply(Complete_metrics, as.character), stringsAsFactors = FALSE)
Nohub_metrics <- data.frame(lapply(Nohub_metrics, as.character), stringsAsFactors = FALSE)
Nobasid_metrics <- data.frame(lapply(Nobasid_metrics, as.character), stringsAsFactors = FALSE)


print(igraph::vcount(Complete)) # Number of nodes
print(igraph::ecount(Complete)) # Number of edges

print(igraph::vcount(NoBasid))
print(igraph::ecount(NoBasid))

print(igraph::vcount(NoHubs))
print(igraph::ecount(NoHubs))


metrics_scaled <- bind_rows(
  Complete_metrics %>% mutate(Network = "Complete"),
  Nohub_metrics %>% mutate(Network = "NoHubs"),
  Nobasid_metrics %>% mutate(Network = "NoBasid")
) %>%
  dplyr::mutate(dplyr::across(where(is.numeric), scale))

metrics_long_scaled <- metrics_scaled %>%
  tidyr::pivot_longer(cols = -c(Node, Network), names_to = "Metric", values_to = "Value")
```



bind the metrics to plot them


```{r}


# 3. Visualization


# Remove missing values
metrics_long_scaled <- na.omit(metrics_long_scaled)

# We visualize only six metrics
selected_metrics <- c(
  "Degree", "Closeness", "Betweenness",
  "EigenvectorCentrality", "PageRank", "Transitivity"
)

metrics_long_filtered <- metrics_long_scaled %>%
  filter(Metric %in% selected_metrics) %>%
  mutate(
    Value = as.numeric(as.character(Value)),
    Network = recode(Network,
      "Complete" = "Complete Network",
      "NoHubs" = "Network & Module Hubs Removed",
      "NoBasid" = "Basidiobolus Subnetwork Removed" ) ) %>%
  na.omit() # Remove any NA if any

metrics_long_filtered$Network <- factor(metrics_long_filtered$Network,
  levels = c(
    "Complete Network",
    "Network & Module Hubs Removed",
    "Basidiobolus Subnetwork Removed" ))

# DspikeIn::color_palette$light_MG
network_colors <- c(
  "Complete Network" = "#F1E0C5",
  "Network & Module Hubs Removed" = "#D2A5A1",
  "Basidiobolus Subnetwork Removed" = "#B2C3A8")

# statistical comparisons a vs b
comparisons <- list(
  c("Complete Network", "Network & Module Hubs Removed"),
  c("Complete Network", "Basidiobolus Subnetwork Removed"),
  c("Network & Module Hubs Removed", "Basidiobolus Subnetwork Removed"))

networks_in_data <- unique(metrics_long_filtered$Network)
comparisons <- comparisons[sapply(comparisons, function(pair) all(pair %in% networks_in_data))]

met <- ggplot(metrics_long_filtered, aes(x = Network, y = Value, fill = Network)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(aes(color = Network),
    position = position_jitter(0.2), alpha = 0.2, size = 1.5 ) +
  scale_fill_manual(values = network_colors) +
  scale_color_manual(values = network_colors) +
  facet_wrap(~Metric, scales = "free_y", labeller = label_wrap_gen(width = 20)) +
  ggpubr::stat_compare_means(method = "wilcox.test", label = "p.signif", comparisons = comparisons) +
  theme_minimal() +
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.x = element_text(size = 10, angle = 10, hjust = 0.8),
    strip.text = element_text(size = 12, margin = margin(t = 15, b = 5)),
    legend.position = "top",
    legend.text = element_text(size = 13),
    legend.title = element_text(size = 13, face = "bold"),
    plot.title = element_text(size = 13, face = "bold")
  ) + labs(title = "Selected Node Metrics Across Networks", fill = "Network Type", color = "Network Type")
```


#          To find first and second neighbors your target node  


```{r}
Complete <- load_graphml("Complete.graphml")
result2 <- extract_neighbors(
  graph = Complete,
  target_node = "OTU69:Basidiobolus_sp", mode = "all")
head(result2$summary)
```



#                 Compute Node-Level Metrics

 | **Metric**                     | **Description** |
 |--------------------------------|-----------------------------------------------------------|
 | 'Node'                         | Node name (character format) |
 | 'Degree'                       | Number of edges connected to the node |
 | 'Strength'                     | Sum of edge weights connected to the node |
 | 'Closeness'                    | Closeness centrality (normalized, based on shortest paths) |
 | 'Betweenness'                  | Betweenness centrality (normalized, measures control over network flow) |
 | 'EigenvectorCentrality'        | Eigenvector centrality (importance based on connections to influential nodes) |
 | 'PageRank'                     | PageRank score (importance based on incoming links) |
 | 'Transitivity'                 | Local clustering coefficient (tendency of a node to form triangles) |
 | 'Coreness'                     | Node's coreness (from k-core decomposition) |
 | 'Constraint'                   | Burt's constraint (measures structural holes in a node's ego network) |
 | 'EffectiveSize'                | Inverse of constraint (larger values = more non-redundant connections) |
 | 'Redundancy'                   | Sum of constraint values of a node's alters |
 | 'Community'                    | Community assignment from Louvain clustering |
 | 'Efficiency'                   | Global efficiency (average inverse shortest path length) |
 | 'Local_Efficiency'             | Local efficiency (subgraph efficiency for a node's neighbors) |
 | 'Within_Module_Connectivity'   | Proportion of neighbors in the same community |
 | 'Among_Module_Connectivity'    | Proportion of neighbors in different communities |
 
 
```{r}
# Load required libraries
library(igraph)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggrepel)

# Compute Node-Level Metrics
completeMetrics <- node_level_metrics(Complete)
NoHubsMetrics <- node_level_metrics(NoHubs)
NoBasidMetrics <- node_level_metrics(NoBasid)

# completeMetrics$facet_plot

# Ensure each dataset has a "Network" column before combining
completeMetrics$metrics <- completeMetrics$metrics %>% mutate(Network = "Complete Network")
NoHubsMetrics$metrics <- NoHubsMetrics$metrics %>% mutate(Network = "Network & Module Hubs Removed")
NoBasidMetrics$metrics <- NoBasidMetrics$metrics %>% mutate(Network = "Basidiobolus Subnetwork Removed")

# Combine All Data
combined_data <- bind_rows(
  completeMetrics$metrics,
  NoHubsMetrics$metrics,
  NoBasidMetrics$metrics
)


# Add Node Identifier if missing
if (!"Node" %in% colnames(combined_data)) {
  combined_data <- combined_data %>% mutate(Node = rownames(.))
}

# Convert `Network` into Factor
combined_data$Network <- factor(combined_data$Network, levels = c(
  "Complete Network",
  "Network & Module Hubs Removed",
  "Basidiobolus Subnetwork Removed"))

# Convert Data to Long Format
metrics_long <- combined_data %>%
  pivot_longer(
    cols = c("Redundancy", "Efficiency", "Betweenness"),
    names_to = "Metric", values_to = "Value")


# Define Custom Colors and Shapes
  network_colors <- c(
  "Complete Network" = "#F1E0C5",
  "Network & Module Hubs Removed" = "#D2A5A1",
  "Basidiobolus Subnetwork Removed" = "#B2C3A8")
  
network_shapes <- c(
  "Complete Network" = 21,
  "Network & Module Hubs Removed" = 22,
  "Basidiobolus Subnetwork Removed" = 23)

# Determine Top 30% of Nodes to Label/Optional
metrics_long <- metrics_long %>%
  group_by(Network, Metric) %>%
  mutate(Label = ifelse(rank(-Value, ties.method = "random") / n() <= 0.3, Node, NA))

# ?quadrant_plot() can creat plot for indivisual network
# plot <- quadrant_plot(metrics, x_metric = "Degree", y_metric = "Efficiency")

# Create comparision Plots
create_metric_plot <- function(metric_name, data, title) {
  data_filtered <- data %>% filter(Metric == metric_name)
  median_degree <- median(data_filtered$Degree, na.rm = TRUE)
  median_value <- median(data_filtered$Value, na.rm = TRUE)

  ggplot(data_filtered, aes(x = Degree, y = Value, fill = Network)) +
    geom_point(aes(shape = Network), size = 3, stroke = 1, color = "black") +
    geom_text_repel(aes(label = Label), size = 3, max.overlaps = 50) +
    scale_fill_manual(values = network_colors) +
    scale_shape_manual(values = network_shapes) +
    geom_vline(xintercept = median_degree, linetype = "dashed", color = "black", size = 1) +
    geom_hline(yintercept = median_value, linetype = "dashed", color = "black", size = 1) +
    labs(
      title = title,
      x = "Degree",
      y = metric_name,
      fill = "Network",
      shape = "Network" ) +
    theme_minimal() +
    theme(
      plot.title = element_text(
        hjust = 0.5, size = 16, face = "bold",
        margin = margin(t = 10, b = 20) # Moves the title downward
      ),
      axis.title = element_text(size = 14, face = "bold"),
      legend.position = "top",
      legend.title = element_text(size = 14, face = "bold"),
      legend.text = element_text(size = 12)  )
}


# Generate Plots
plot_redundancy <- create_metric_plot("Redundancy", metrics_long, "Redundancy vs. Degree Across Networks")
plot_efficiency <- create_metric_plot("Efficiency", metrics_long, "Efficiency vs. Degree Across Networks")
plot_betweenness <- create_metric_plot("Betweenness", metrics_long, "Betweenness vs. Degree Across Networks")

# Save Plots
# ggsave("plot_redundancy_20percent.png", plot_redundancy, width = 8, height = 6)
# ggsave("plot_efficiency_20percent.png", plot_efficiency, width = 8, height = 6)
# ggsave("plot_betweenness_20percent.png", plot_betweenness, width = 8, height = 6)

# Print Plots
# print(plot_redundancy)
# print(plot_efficiency)
# print(plot_betweenness)
```



#            Compute degree metrics and visualize the network


```{r}
# Compute degree metrics and visualize the network
# Options: `"stress"` (default), `"graphopt"`, `"fr"`

result <- degree_network(graph_path = Complete, save_metrics = TRUE)
# print(result$metrics)
# print(result$plot)


# Compute network weights for different graph structures
NH <- weight_Network(graph_path = "NoHubs.graphml")
NB <- weight_Network(graph_path = "NoBasid.graphml")
C <- weight_Network(graph_path = "Complete.graphml")

# Extract metrics from the computed network weights
CompleteM <- C$metrics
NoHubsM <- NH$metrics
NoBasidM <- NB$metrics

# Combine metrics into a single dataframe for comparison
df <- bind_rows(
  CompleteM %>% mutate(Group = "CompleteM"),
  NoHubsM %>% mutate(Group = "NoHubsM"),
  NoBasidM %>% mutate(Group = "NoBasidM")) %>%
  pivot_longer(cols = -Group, names_to = "Metric", values_to = "Value")

# Aggregate the total values by metric and group
df_bar <- df %>%
  group_by(Metric, Group) %>%
  summarise(Total_Value = sum(Value), .groups = "drop")

# Plot the metrics comparison
pg<-ggplot(df_bar, aes(x = Metric, y = log1p(Total_Value), fill = Group)) +
  geom_bar(stat = "identity", position = "dodge", alpha = 0.8) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_manual(values = c("#F1E0C5", "#D2A5A1", "#B2C3A8")) +
  labs(
    title = "Total Network Metrics Comparison",
    x = "Metric",
    y = "Total Value (log-scaled)",
    fill = "Group" )
```

# Small-World Index Analysis 
To determine whether the complete network exhibited small-world topology, we computed the Small-World Index (SWI; σ) following the quantitative framework established by Humphries M, Gurney K, 2008. PLoS ONE 4
(Eq.1) (SWI; σ) = (Global clustering coefficient real/Global clustering coefficient random)/(Avg Path real/Avg Path random))

```{r, eval=FALSE}
library(igraph)
library(tidygraph)
library(ggraph)
library(DspikeIn)


# AA abundance
Complete<-load_graphml("Complete.graphml")
# NoHubs<-load_graphml("NoHubs.graphml")
# NoBasid<-load_graphml("NoBasid.graphml")

# RA abundance
# Complete_Rel<-load_graphml("~/Downloads/herp.spiecsym.Rel.graphml")

# Degree distribution
# deg <- degree(Complete_Rel)
deg <- degree(Complete)

hist(deg, breaks = 30, main = "Degree Distribution", xlab = "Degree")
fit <- fit_power_law(deg + 1)  # Avoid zero degrees
print(fit)


# ---- empirical network ----
g_empirical = Complete
g_empirical = Complete_Rel



# ---- Calculate real network metrics ----
creal <- transitivity(g_empirical, type = "global")
E(g_empirical)$weight <- abs(E(g_empirical)$weight)
lreal <- mean_distance(g_empirical, directed = FALSE, unconnected = TRUE, weights = E(g_empirical)$weight)

# ---- Generate 1000 Erdős-Rényi random graphs with same size ----
set.seed(42)  # for reproducibility
n_nodes <- vcount(g_empirical)
n_edges <- ecount(g_empirical)

crand_vals <- numeric(1000)
lrand_vals <- numeric(1000)

for (i in 1:1000) {
g_rand <- sample_gnm(n = n_nodes, m = n_edges, directed = FALSE)
if (!is_connected(g_rand)) next
  
  crand_vals[i] <- transitivity(g_rand, type = "global")
  lrand_vals[i] <- mean_distance(g_rand, directed = FALSE, unconnected = TRUE)
}

# ---- Calculate mean values across random graphs ----
crand <- mean(crand_vals, na.rm = TRUE)
lrand <- mean(lrand_vals, na.rm = TRUE)

# ---- Compute Small-World Index (σ) ----
sigma <- (creal / crand) / (lreal / lrand)


cat("Global clustering coefficient (real):", creal, "\n")
cat("Average path length (real):", lreal, "\n")
cat("Mean clustering coefficient (random):", crand, "\n")
cat("Mean path length (random):", lrand, "\n")
cat("Small-World Index (σ):", round(sigma, 2), "\n")

```


```{r}
sessionInfo()
```


```{r cleanup, include = FALSE}
# =====================================================================
# FINAL CLEANUP: Remove files created during vignette execution
# =====================================================================

# **Note:** To comply with CRAN policies, this vignette removes any output files
# (e.g., plots, summary tables) created during execution.
# You may modify paths locally if you'd like to keep the outputs.

# List of all files possibly written by example functions
files_to_delete <- c(
  "Global_Network_Metrics.csv", "absolute_counts.csv",
  "abundance_analysis_filtered_phyloseq.rds", "abundance_analysis_filtered_tse.rds",
  "abundance_analysis_high_abundance_phyloseq.rds", "abundance_analysis_high_abundance_tse.rds",
  "abundance_analysis_low_abundance_phyloseq.rds", "abundance_analysis_low_abundance_tse.rds",
  "adjusted_prevalence.rds", "core_microbiome.csv", "core_microbiome.rds", "comp_spikein_patristic_distances.csv", "comp_spikein_branch_lengths.pdf", "comp_spikein_patristic_distances_hist.pdf",
  "merged_data.csv", "merged_data.docx", "comp_spikein.nwk", "comp_spikein_tree.png", "comp_spikein_patristic_distances_hist.pfd","post_eval_summary.csv","post_eval_summary.docx",
  "merged_physeq_max_processed.rds", "merged_physeq_sum.rds",
  "merged_physeq_sum_processed.rds", "spikeIn_factors_output",
  "plot_Spike.percentage_Kruskal.pdf", "plot_Spike.percentage_Kruskal.png",
  "plot_Spike.reads_Kruskal.pdf", "plot_Spike.reads_Kruskal.png",
  "plot_Total.reads_Kruskal.pdf", "plot_Total.reads_Kruskal.png",
  "proportion_adjusted_physeq.rds", "proportion_adjusted_tse.rds",
  "randomsubsample_Trimmed_evenDepth.rds", "summary.csv", "summary.docx", "DspikeIn-manual.tex", "output.csv", "output.docx",
  "tree_with_bootstrap_and_cophenetic.png"
)

# Remove files safely
invisible(file.remove(files_to_delete[file.exists(files_to_delete)]))

# Remove known folders (if created by functions)
folders_to_delete <- c("spikeIn_factors_output")
for (folder in folders_to_delete) {
  if (dir.exists(folder)) {
    unlink(folder, recursive = TRUE, force = TRUE)
  }
}
```


#         DspikeIn volume protocol

Spike-in volume Protocol; 

The species Tetragenococcus halophilus (bacterial spike; ATCC33315) and Dekkera bruxellensis (fungal spike; WLP4642-White Labs) were selected as taxa to spike into gut microbiome samples as they were not found in an extensive collection of wildlife skin (GenBank BioProjects: PRJNA1114724, PRJNA 1114659) or gut microbiome samples. Stock cell suspensions of both microbes were grown in either static tryptic soy broth (T. halophilus) or potato dextrose broth (D. bruxellensis) for 72 hours then serially diluted and optical density (OD600) determined on a ClarioStar plate reader. Cell suspensions with an optical density of 1.0, 0.1, 0.01, 0.001 were DNA extracted using the Qiagen DNeasy Powersoil Pro Kit. These DNA isolations were used as standards to determine the proper spike in volume of cells to represent 0.1-10% of a sample (Rao et al., 2021b) Fecal pellets (3.1 ± 1.6 mg; range = 1 – 5.1 mg) from an ongoing live animal study using wood frogs (Lithobates sylvaticus) were used to standardize the input material for the development of this protocol. A total of (n=9) samples were used to validate the spike in protocol. Each fecal sample was homogenized in 1mL of sterile molecular grade water then 250uL of fecal slurry was DNA extracted as above with and without spiked cells. 
Two approaches were used to evaluate the target spike-in of 0.1-10%, the range of effective spike-in percentage described in (Rao et al., 2021b), including 1) an expected increase of qPCR cycle threshold (Ct) value that is proportional to the amount of spiked cells and 2) the expected increase in copy number of T. halophilus and D. bruxellensis in spiked vs. unspiked samples. A standard curve was generated using a synthetic fragment of DNA for the 16S-V4 rRNA and ITS1 rDNA regions of T. halophilus and D. bruxellensis, respectively. The standard curve was used to convert Ct values into log copy number for statistical analyses (detailed approach in[2, 3]) using the formula y = -0.2426x + 10.584 for T. halophilus and y = -0.3071x + 10.349 for D. bruxellensis, where x is the average Ct for each unknown sample.
Quantitative PCR (qPCR) was used to compare known copy numbers from synthetic DNA sequences of T. halophilus and D. bruxellensis to DNA extractions of T. halophilus and D. bruxellensis independently, and wood frog fecal samples with and without spiked cells. SYBR qPCR assays were run at 20ul total volume including 10ul 2X Quantabio PerfeCTa SYBR Green Fastmix, 1ul of 10uM forward and reverse primers, 1ul of ArcticEnzymes dsDNAse master mix clean up kit, and either 1ul of DNA for D. bruxellensis or 3ul for T. halophilus. Different volumes of DNA were chosen for amplification of bacteria and fungi due to previous optimization of library preparation and sequencing steps [3]. The 515F [4] and 806R [5] primers were chosen to amplify bacteria and ITS1FI2 [6] and ITS2 for fungi, as these are the same primers used during amplicon library preparation and sequencing. Cycling conditions on an Agilent AriaMX consisted of 95 C for 3 mins followed by 40 cycles of 95 C for 15 sec, 60 C for 30 sec and 72 C for 30 sec. Following amplification, a melt curve was generated under the following conditions including 95 C for 30 sec, and a melt from 60 C to 90 C increasing in resolution of 0.5 C in increments of a 5 sec soak time. To validate the spike in protocol we selected two sets of fecal samples including 360 samples from a diverse species pool of frogs, lizards, salamanders and snakes and a more targeted approach of 122 fecal samples from three genera of salamanders from the Plethodontidae. (Supplemental Table #). FFecal samples were not weighed in the field, rather, a complete fecal pellet was diluted in an equal volume of sterile water and standardized volume of fecal slurry (250µL) extracted for independent samples.. A volume of 1ul T. halophilus (1847 copies) and 1ul D. bruxellensis (733 copies) were spiked into each fecal sample then DNA was extracted as above, libraries constructed, and amplicon sequenced on an Illumina MiSeq as in [7] .

<https://github.com/mghotbi/DspikeIn>
