---
title: "Getting started with DspikeIn"
author: "Mitra Ghotbi"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
    highlight: tango
    df_print: tibble
    css: "style.css"
vignette: >
  %\VignetteIndexEntry{Getting started with DspikeIn}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}


---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```



```{r figure_setup, include=FALSE}
knitr::opts_chunk$set(
  fig.width = 8,
  fig.height = 5,
  fig.align = "center",
  dpi = 300,
  out.width = "60%",
  cache = TRUE,
  dev = "png"
)
```


#                Required Packages


```{r, eval=FALSE}
# ============================================================================
#                         INSTALL CRAN PACKAGES
# ============================================================================

# Install missing CRAN packages
install.packages(setdiff(
  c(
    "stats", "dplyr", "ggplot2", "flextable", "ggpubr",
    "randomForest", "ggridges", "ggalluvial", "tibble",
    "matrixStats", "RColorBrewer", "ape", "rlang",
    "scales", "magrittr", "phangorn", "igraph", "tidyr",
    "xml2", "data.table", "reshape2", "vegan", "patchwork", "officer"
  ),
  installed.packages()[, "Package"]
))

# Load CRAN packages
lapply(c(
  "stats", "dplyr", "ggplot2", "flextable", "ggpubr", "randomForest",
  "ggridges", "ggalluvial", "tibble", "matrixStats", "RColorBrewer",
  "ape", "rlang", "scales", "magrittr", "phangorn", "igraph", "tidyr",
  "xml2", "data.table", "reshape2", "vegan", "patchwork", "officer"
), library, character.only = TRUE)

# ============================================================================
#                         INSTALL BIOCONDUCTOR PACKAGES
# ============================================================================

# Install BiocManager if not installed
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")

# Install missing Bioconductor packages
BiocManager::install(setdiff(
  c(
    "phyloseq", "msa", "DESeq2", "ggtree", "edgeR",
    "Biostrings", "DECIPHER", "microbiome", "limma",
    "S4Vectors", "SummarizedExperiment", "TreeSummarizedExperiment"
  ),
  installed.packages()[, "Package"]
))

# Load Bioconductor packages
lapply(
  c(
    "phyloseq", "msa", "DESeq2", "edgeR", "Biostrings", "ggtree", "DECIPHER",
    "microbiome", "limma", "S4Vectors", "SummarizedExperiment", "TreeSummarizedExperiment"
  ),
  library,
  character.only = TRUE
)

# ============================================================================
#                         INSTALL DspikeIn FROM GITHUB
# ============================================================================

# To access the DspikeIn vignette for a detailed tutorial, use vignette("DspikeIn"), or browse all available vignettes with browseVignettes("DspikeIn").
devtools::install_github("mghotbi/DspikeIn", build_vignettes = TRUE, dependencies = TRUE)
library(DspikeIn)
browseVignettes("DspikeIn")
vignette("DspikeIn")

## or

if (!requireNamespace("devtools", quietly = TRUE)) install.packages("devtools")
devtools::install_github("mghotbi/DspikeIn")

# Load DspikeIn only if installed
if ("DspikeIn" %in% installed.packages()[, "Package"]) {
  library(DspikeIn)
} else {
  stop("DspikeIn installation failed. Check errors above.")
}
```


#    Acknowledgments
The development of the DspikeIn package was made possible through the invaluable contributions of the R and Bioconductor communities, whose open-source tools provided the foundation for this work. We are especially grateful to the developers of phyloseq, DESeq2, edgeR, ggplot2, ggtree, TreeSummarizedExperiment,vegan, ggtreeExtra and other essential packages that enable robust analysis and visualization of microbiome data. We further acknowledge the broader scientific community working on absolute microbial quantification and compositional data analysis, whose pioneering research inspired the design of DspikeIn.


#     DspikeIn
The DspikeIn package supports both phyloseq and TreeSummarizedExperiment formats to streamline microbial quantification across diverse experimental setups. It accommodates either a single spike-in taxon or synthetic community taxa with variable or equal spike-in volumes and copy numbers. The package offers a comprehensive suite of tools for AA quantification, addressing challenges through ten core functions: 1) validation of spiked species, 2) data preprocessing, 3) system-specific spiked species retrieval, 4) scaling factor calculation, 5) conversion to absolute abundance, 6) bias correction and normalization, 7) performance assessment, and 8) taxa exploration and filtering 9) network topology assessment 10) further analyses and visualization. 


#    DspikeIn requirements
DspikeIn works with 7 taxonomic ranks


```{r, eval=FALSE}
# =====================================================================
#             To remove strain from the taxonomic ranks
# =====================================================================
# DspikeIn works with 7 taxonomic ranks
# colnames(taxmat) <- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")

library(phyloseq)
# Function to remove strain information from taxonomy columns
remove_strain_info <- function(tax_table) {
  # Define the regex pattern for common strain identifiers
  pattern <- "Strain.*|strain.*|\\s*\\[.*\\]|\\s*\\(.*\\)" # Adjust the regex to match specific strain formats
  # Apply the pattern to each column of the taxonomy table
  for (col in colnames(tax_table)) {
    tax_table[, col] <- gsub(pattern, "", tax_table[, col]) # Remove strain info
    tax_table[, col] <- trimws(tax_table[, col]) # Trim trailing whitespace
  }
  return(tax_table)
}
# Step 1: Extract the taxonomy table
taxonomy <- tax_table(ps)
# Step 2: Remove strain information (including the `Strain` column)
cleaned_taxonomy <- remove_strain_info(taxonomy)
# Remove the `Strain` column if it exists
if ("Strain" %in% colnames(cleaned_taxonomy)) {
  cleaned_taxonomy <- cleaned_taxonomy[, colnames(cleaned_taxonomy) != "Strain"]
}
# Step 3: Update the taxonomy table in the phyloseq object
tax_table(ps) <- cleaned_taxonomy
# Step 4: Verify the changes
print(head(tax_table(ps))) # Display the first few rows


# =====================================================================
#             To add species rank to the taxonomic ranks
# =====================================================================

library(phyloseq)

# Step 1: Extract taxonomy table safely
taxonomy <- as.data.frame(as.matrix(tax_table(ps)))
if (!"Genus" %in% colnames(taxonomy)) {
  stop("The 'Genus' column is missing in the taxonomy table.")
}

# Step 2: Handle potential missing genera (optional but recommended)
taxonomy$Genus[is.na(taxonomy$Genus) | taxonomy$Genus == ""] <- "Unknown"
# Step 3: Create a new 'species' column
taxonomy$species <- paste0(taxonomy$Genus, "_OTU", seq_len(nrow(taxonomy)))
# Step 4: Assign back to phyloseq object (must coerce back to matrix)
tax_table(ps) <- tax_table(as.matrix(taxonomy))
```



#        Build phyloseq or TreeSummarizedExperiment file
for more information please refer to https://github.com/joey711/phyloseq & 
https://github.com/markrobinsonuzh/TreeSummarizedExperiment


```{r, eval=FALSE}
# =====================================================================
#                     Build phyloseq
# =====================================================================
otu <- read.csv("otu.csv", header = TRUE, sep = ",", row.names = 1)
# taxonomic rank need to be capilalized, only the first letter of each rank
tax <- read.csv("tax.csv", header = TRUE, sep = ",", row.names = 1)
# Ensure 'spiked.volume' column is present and correctly formatted in metadata
meta <- read.csv("metadata.csv", header = TRUE, sep = ",")

# Convert data to appropriate formats
meta <- as.data.frame(meta)
taxmat <- as.matrix(tax)
otumat <- as.matrix(otu)
colnames(taxmat) <- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")
OTU <- otu_table(otumat, taxa_are_rows = TRUE)
TAX <- phyloseq::tax_table(taxmat)

# Check
row.names(meta) <- sample_names(OTU)
metadata <- sample_data(meta)
# Build phyloseq obj
physeq <- phyloseq(OTU, TAX, metadata)

# Follow the next steps if tree and reference files are included
MyTree <- read.tree("tree.nwk")
reference_seqs <- readDNAStringSet(file = "dna-sequences.fasta", format = "fasta")

physeq_16SOTU <- merge_phyloseq(physeq, reference_seqs, MyTree)
physeq_16SOTU <- tidy_phyloseq_tse(physeq_16SOTU)

saveRDS(physeq_16SOTU, file = "physeq_16SOTU.rds")
physeq_16SOTU <- readRDS("physeq_16SOTU.rds")


# =====================================================================
#                       Build TreeSummarizedExperiment (TSE)
# =====================================================================

otu <- read.csv("otu.csv", header = TRUE, sep = ",", row.names = 1)
otu_mat <- as.matrix(otu) # Convert to matrix
tax <- read.csv("tax.csv", header = TRUE, sep = ",", row.names = 1)
colnames(tax) <- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")
tax_mat <- as.matrix(tax) # Convert to matrix
meta <- read.csv("metadata.csv", header = TRUE, sep = ",", row.names = 1)
reference_seqs <- readDNAStringSet("dna-sequences.fasta", format = "fasta")
tse <- TreeSummarizedExperiment(
  assays = list(counts = otu_mat), # OTU table
  rowData = tax_mat, # Taxonomy information
  colData = meta, # Sample metadata
  rowTree = MyTree, # Phylogenetic tree
  rowSeqs = reference_seqs # Reference sequences
)
```



Do all detected sample spike-in sequences cluster with the reference, and are their branch lengths statistically similar, supporting a common ancestor?

# spike-in validation 

All sample-derived sequences are forming a clade with the reference.
We look for a monophyletic grouping of spike-in OTUs
The clade is strongly supported (bootstrap around 100 percentage).
The branch lengths and distances are in a biologically plausible range.


```{r}
# Use the Neighbor-Joining method  based on a Jukes-Cantor distance matrix and plot the tree with bootstrap values.
# we compare the Sanger read of Tetragenococcus halophilus with the FASTA sequence of Tetragenococcus halophilus from our phyloseq object.

library(Biostrings)
library(phyloseq)
library(DspikeIn)

# Get path to external data folder
extdata_path <- system.file("extdata", package = "DspikeIn")
list.files(extdata_path)

data("physeq_16SOTU", package = "DspikeIn")

physeq_16SOTU <- tidy_phyloseq_tse(physeq_16SOTU)

physeq_16SOTU <- phyloseq::prune_taxa(
  get_tax_table(physeq_16SOTU)$Kingdom %in% c("Bacteria", "Archaea"),
  physeq_16SOTU
)


# Subset the phyloseq object to include only Tetragenococcus species first
# Tetragenococcus <- subset_taxa(physeq_16SOTU, Genus=="Tetragenococcus")
# Tetragenococcus <- subset_taxa(Tetragenococcus, !is.na(taxa_names(Tetragenococcus)) & taxa_names(Tetragenococcus) != "")
# tree <- phy_tree(Tetragenococcus)
# ref_sequences_Tetragenococcus <- refseq(Tetragenococcus)
# library(Biostrings)
# writeXStringSet(ref_sequences_Tetragenococcus, "Sample.fasta.fasta")



ref_fasta <- system.file("extdata", "Ref.fasta", package = "DspikeIn")
sample_fasta <- system.file("extdata", "Sample.fasta", package = "DspikeIn")



result <- validate_spikein_clade(
  reference_fasta = ref_fasta,
  sample_fasta = sample_fasta,
  bootstrap = 200,
  output_prefix = NULL
)

result$tree_plot
```


#     Did spike-ins behave as expected across all samples?
Tip labels=	OTU/ASV names
Branch length numbers=	Actual evolutionary distances (small = very similar)
Prevalence stars	How frequently the OTU occurs across samples
Blue bar ring=	Log10 mean abundance
Outer colored tiles=	The metadata variable you choose (e.g., Animal.type)


```{r}
data("physeq_16SOTU", package = "DspikeIn")
library(ggstar)
library(ggplot2)
# filter your object to only include spike-in taxa beforehand:
# change the OTU IDs for easy detection
# Big stars = detected in many samples
# Small stars = rarely detected
# log10(Mean Abundance) Bars= Color intensity reflects mean abundance.
# The log-transformed average abundance of each OTU across all samples where it appears.
# Extreme blue may signal unintended over-representation.
# Metadata Ring = factor of your interest e.g. Animal.type
# Each OTU is colored by where it was observed.
# Branch length numbers=	Actual evolutionary distances (small = very similar)


spikein <- phyloseq::subset_taxa(physeq_16SOTU, Genus == "Tetragenococcus")
taxa_names(spikein) <- paste0("OTU", seq_len(ntaxa(spikein)))

# Visualize
ps <- plot_spikein_tree_diagnostic(
  obj = spikein,
  metadata_var = "Animal.type",
  output_prefix = "tetragenococcus_diag"
)

ps
```


#                Pre_processing
merges monophyletic ASVs/OTUs

```{r}
# merges monophyletic ASVs/OTUs

# The function Pre_processing_species() merges ASVs/OTUs
# of a species using "sum" or "max" methods, preserving taxonomic,
# phylogenetic, and sequencing data.

# =====================================================================
#                      Load the phyloseq objs/ TSE obj
# =====================================================================

library(phyloseq)
library(DspikeIn)
library(TreeSummarizedExperiment)
library(SummarizedExperiment)

data("physeq_16SOTU", package = "DspikeIn")

# TSE format
# tse_16SOTU <- convert_phyloseq_to_tse(physeq_16SOTU)
# physeq_16SOTU <- convert_tse_to_phyloseq(tse_16SOTU)

physeq_16SOTU <- DspikeIn::tidy_phyloseq_tse(physeq_16SOTU) # make it tidy

# Check if metadata contains spiked volumes with this format
colnames(sample_data(physeq_16SOTU))
```


#          Spiked species and related parameters for 16S


```{r}
# =====================================================================
#                      PREREQUISITE FOR 16S & CALCULATE SPIKED %
# =====================================================================
# Define spiked species and related parameters**

library(flextable)
library(DspikeIn)
# Define spike-in parameters
spiked_cells <- 1847
species_name <- spiked_species <- c("Tetragenococcus_halophilus", "Tetragenococcus_sp.")
merged_spiked_species <- "Tetragenococcus_halophilus"

# If you prefer Genus-level matching
# species_name <- spiked_species <- c("Tetragenococcus")
# merged_spiked_species <- "Tetragenococcus"


# Subset taxa for spiked species
Tetragenococcus <- phyloseq::subset_taxa(
  physeq_16SOTU,
  Species %in% species_name
)

# Get taxon hashcodes
hashcodes <- row.names(phyloseq::tax_table(Tetragenococcus))

# Subset samples based on spiked volume
spiked_16S_OTU_spiked <- phyloseq::subset_samples(
  physeq_16SOTU,
  spiked.volume %in% c("2", "1")
)

# Merge OTUs derived from spiked species
# File will be saved to tempdir() (no cleanup needed)
output_rds <- file.path(tempdir(), "merged_physeq_sum.rds")

Spiked_16S_sum_scaled <- Pre_processing_species(
  spiked_16S_OTU_spiked,
  species_name,
  merge_method = "sum",
  output_file = output_rds
)

# Calculate the spike-in percentage across samples
Perc <- calculate_spike_percentage(
  Spiked_16S_sum_scaled,
  merged_spiked_species,
  passed_range = c(0.1, 20)
)
```



#          Spiked species and related parameters for ITS



```{r, eval=FALSE}
# =====================================================================
#                      PREREQUISITE FOR ITS & CALCULATE SPIKED %
# =====================================================================
# Define spiked species and related parameters**


# Define the spiked species
# spiked_cells <- 733
# species_name <- spiked_species <- merged_spiked_species <- "Dekkera_bruxellensis"

# Subset taxa for spiked species
# Dekkera <- phyloseq::subset_taxa(
# physeq_ITSOTU,
# Species %in% species_name)

# hashcodes <- row.names(phyloseq::tax_table(Dekkera))

# Subset samples based on spiked volume
# physeq_ITSOTU_spiked <- phyloseq::subset_samples(physeq_ITSOTU, spiked.volume %in% c("2", "1"))

# if TSE format
# tse_ITSOTU <- convert_phyloseq_to_tse(physeq_ITSOTU)
# physeq_ITSOTU_spiked <- tse_ITSOTU[, tse_ITSOTU$spiked.volume %in% c("2", "1")]
```




#         Calculating Scaling Factors


```{r}
#
# =====================================================================
#                      CALCULATE SCALING FACTORS
# =====================================================================


# Calculate spike-in scaling factors
result <- calculate_spikeIn_factors(
  obj = Spiked_16S_sum_scaled,
  spiked_cells = spiked_cells,
  merged_spiked_species = merged_spiked_species
)

# View extracted outputs
result$spiked_species_merged # Merged spiked species name
result$spiked_species_reads # Total reads detected for the spike
scaling_factors <- result$scaling_factors
head(scaling_factors) # Vector of scaling factors per sample
```



#                Convert relative counts to absolute counts


```{r}
# =====================================================================
#             Convert relative counts to absolute counts
# =====================================================================

# absolute counts=relative counts×sample-specific scaling factor


# Convert to absolute counts
absolute <- convert_to_absolute_counts(Spiked_16S_sum_scaled, scaling_factors)

# Extract processed data
absolute_counts <- absolute$absolute_counts
physeq_absolute <- absolute$obj_adj

physeq_absolute <- tidy_phyloseq_tse(physeq_absolute)

# View absolute count data
head(absolute_counts)
```


#     Summary Stat


```{r}
# =====================================================================
#                      CALCULATE SPIKE PERCENTAGE & summary stat
# =====================================================================

# ** Calculate spike percentage & Generate summary statistics for absolute counts**

# Generate summary statistics for absolute counts
post_eval_summary <- calculate_summary_stats_table(absolute_counts)

# You may want to Back normal to check calculation accuracy
# the scaling factor was computed based on spiked species reads and fixed cell count.
# Multiplying the spiked species read count by this scaling factor restores the exact spiked cell count.
# lets check it
# BackNormal <- calculate_spike_percentage(
#  physeq_absolute,
#  merged_spiked_species,
#  passed_range = c(0.1, 20)
# )
```



#        Filter unsuccessful spiked samples


```{r}
# Optional, or you may do it at the end of the process
# ** Time to filter out unsuccessful spiked samples**

library(phyloseq)
library(dplyr)
library(tibble)
library(microbiome)

filtered_sample_data <- microbiome::meta(physeq_absolute) %>%
  as.data.frame() %>%
  tibble::rownames_to_column(var = "Sample") %>%
  dplyr::mutate(Sample = as.character(Sample)) %>%
  dplyr::left_join(Perc, by = "Sample")

filtered_sample_data <- tibble::column_to_rownames(filtered_sample_data, "Sample")

filtered_sample_data <- sample_data(as.data.frame(filtered_sample_data))

# Assign back to phyloseq obj
sample_data(physeq_absolute) <- filtered_sample_data
```


#           spiked species retrieval is system-dependent
The goal is to identify the range where, for example, the evenness of your community remains independent of spiked species retrieval—meaning the p-value should not be significant, and the R² value should be low, indicating minimal influence.


```{r}
#  The acceptable range of spiked species retrieval is system-dependent
# Spiked species become centroid of the community (Distance to Centroid)
# Spiked species become dominant and imbalance the community (Evenness)

# What range of spiked species retrieval is appropriate for your system?

# Calculate Pielou's Evenness using Shannon index and species richness (Observed)
# Load required libraries
library(vegan)
# physeq_ab <- normalization_set(physeq_absolute, method = "rar")
# physeq_absolute <- physeq_ab$dat.normed

# Calculate Pielou's Evenness using Shannon index and species richness (Observed)
alphab <- phyloseq::estimate_richness(physeq_absolute, measures = c("Observed", "Shannon"))
alphab$Pielou_evenness <- alphab$Shannon / alphab$Observed

# Normalize values
alphab <- alphab %>%
  mutate(across(c("Observed", "Shannon", "Pielou_evenness"), ~ as.numeric(scale(.))))

metadata <- as.data.frame(microbiome::meta(physeq_absolute))

metadata$Sample <- rownames(metadata)
alphab$Sample <- rownames(alphab)

# Merge alpha diversity metrics into metadata
metadata <- dplyr::left_join(metadata, alphab[, c("Sample", "Observed", "Shannon", "Pielou_evenness")], by = "Sample")

metadata <- metadata %>%
  column_to_rownames(var = "Sample")

# Updated metadata back to the phyloseq obj
sample_data(physeq_absolute) <- sample_data(metadata)

if (!"Spiked_Reads" %in% colnames(metadata)) {
  stop("Column 'Spiked_Reads' not found in metadata.")
}

# Generate regression plot
plot_object <- regression_plot(
  data = metadata,
  x_var = "Pielou_evenness",
  y_var = "Spiked_Reads",
  custom_range = c(0.1, 20, 30, 50, 100),
  plot_title = NULL
)

plot_object
```



#            Calculate the percentage of spiked species retrieval


```{r}
# * Calculate the percentage of spiked species retrieval per sample*

absolute_abundance_16S_OTU_perc <- phyloseq::subset_samples(physeq_absolute, sample.or.blank != "blank")

# Adjust the threshold range as needed based on system specifications

result_perc <- calculate_spike_percentage(
  absolute_abundance_16S_OTU_perc,
  merged_spiked_species,
  passed_range = c(0.1, 20)
)

# BackNormal
conc <- conclusion(absolute_abundance_16S_OTU_perc,
  merged_spiked_species,
  max_passed_range = 20,
  output_path
)

conc$full_report
conc$summary_stats

# you may keep only passed data
# Filter to get only the samples that passed
passed_samples <- result_perc$Sample[result_perc$Result == "passed"]

# Subset the original phyloseq object to keep only the samples that passed
passed_physeq <- phyloseq::prune_samples(passed_samples, absolute_abundance_16S_OTU_perc)
```



#  Abundance-based Core microbiome


```{r}
physeq_absolute <- absolute$obj_adj
pps_Abs <- get_long_format_data(physeq_absolute)

# calculation for relative abundance needs sum of total reads
# total_reads <- sum(pps_Abs$Abundance)

# Generate an alluvial plot

alluvial_plot_abs <- alluvial_plot(
  data = pps_Abs,
  axes = c("Host.genus", "Ecoregion.III", "Diet"),
  abundance_threshold = 10000,
  fill_variable = "Family",
  silent = TRUE,
  abundance_type = "absolute",
  top_taxa = 15,
  text_size = 4,
  legend_ncol = 1,
  custom_colors = DspikeIn::color_palette$light_MG # Use the color palette from DspikeIn
)

alluvial_plot_abs
```



#                 Data transform & Normalization

you may select to transform your data befor moving forward with Differential Abundance


```{r}
# you may need to normalize/transform your data to reduce biases

ps <- physeq_16SOTU

# TC Normalization
result_TC <- normalization_set(ps, method = "TC", groups = "Host.species")
normalized_ps_TC <- result_TC$dat.normed
scaling_factors_TC <- result_TC$scaling.factor

# UQ Normalization
data("physeq_16SOTU", package = "DspikeIn")
ps <- physeq_16SOTU
result_UQ <- normalization_set(ps, method = "UQ", groups = "Host.species")
normalized_ps_UQ <- result_UQ$dat.normed
scaling_factors_UQ <- result_UQ$scaling.factor

# Median Normalization
data("physeq_16SOTU", package = "DspikeIn")
ps <- physeq_16SOTU
result_med <- normalization_set(ps, method = "med", groups = "Host.species")
normalized_ps_med <- result_med$dat.normed
scaling_factors_med <- result_med$scaling.factor

# DESeq Normalization
data("physeq_16SOTU", package = "DspikeIn")
ps <- physeq_16SOTU
ps_n <- remove_zero_negative_count_samples(ps)
result_DESeq <- normalization_set(ps_n, method = "DESeq", groups = "Animal.type")
normalized_ps_DESeq <- result_DESeq$dat.normed
scaling_factors_DESeq <- result_DESeq$scaling.factor

# Poisson Normalization
data("physeq_16SOTU", package = "DspikeIn")
ps <- physeq_16SOTU
result_Poisson <- normalization_set(ps, method = "Poisson", groups = "Host.genus")
normalized_ps_Poisson <- result_Poisson$dat.normed
scaling_factors_Poisson <- result_Poisson$scaling.factor

# Quantile Normalization
data("physeq_16SOTU", package = "DspikeIn")
ps <- physeq_16SOTU
result_QN <- normalization_set(ps, method = "QN")
normalized_ps_QN <- result_QN$dat.normed
scaling_factors_QN <- result_QN$scaling.factor

# TMM Normalization
data("physeq_16SOTU", package = "DspikeIn")
ps <- physeq_16SOTU
result_TMM <- normalization_set(ps, method = "TMM", groups = "Animal.type")
normalized_ps_TMM <- result_TMM$dat.normed
scaling_factors_TMM <- result_TMM$scaling.factor

# CLR Normalization
data("physeq_16SOTU", package = "DspikeIn")
ps <- physeq_16SOTU
result_clr <- normalization_set(ps, method = "clr")
normalized_ps_clr <- result_clr$dat.normed
scaling_factors_clr <- result_clr$scaling.factor

# Rarefying
data("physeq_16SOTU", package = "DspikeIn")
ps <- physeq_16SOTU
result_rar <- normalization_set(ps, method = "rar")
normalized_ps_rar <- result_rar$dat.normed
scaling_factors_rar <- result_rar$scaling.factor

# CSS Normalization
data("physeq_16SOTU", package = "DspikeIn")
ps <- physeq_16SOTU
result_css <- normalization_set(ps, method = "css")
normalized_ps_css <- result_css$dat.normed
scaling_factors_css <- result_css$scaling.factor

# TSS Normalization
data("physeq_16SOTU", package = "DspikeIn")
ps <- physeq_16SOTU
result_tss <- normalization_set(ps, method = "tss")
normalized_ps_tss <- result_tss$dat.normed
scaling_factors_tss <- result_tss$scaling.factor

# RLE Normalization
data("physeq_16SOTU", package = "DspikeIn")
ps <- physeq_16SOTU
result_rle <- normalization_set(ps, method = "rle")
normalized_ps_rle <- result_rle$dat.normed
scaling_factors_rle <- result_rle$scaling.factor
```




 Ridge plot


```{r}
rf_physeq <- RandomForest_selected(
  physeq_16SOTU,
  response_var = "Host.genus",
  na_vars = c("Habitat", "Ecoregion.III", "Host.genus", "Diet")
)

ridge_physeq <- ridge_plot_it(rf_physeq, taxrank = "Family", top_n = 10) + scale_fill_manual(values = DspikeIn::color_palette$cool_MG)
ridge_physeq

result_css <- normalization_set(rf_physeq, method = "css")
normalized_ps_css <- result_css$dat.normed

ridge_physeq <- ridge_plot_it(normalized_ps_css, taxrank = "Family", top_n = 10) + scale_fill_manual(values = DspikeIn::color_palette$cool_MG)

# ridge_physeq
```



#                  Further analysis 

remove the spike-in sp before further analysis

```{r}
absolute <- phyloseq::subset_taxa(physeq_absolute, Genus != "Tetragenococcus")
absolute <- subset_taxa(absolute, Family != "Chloroplast" & Order != "Chloroplast")
Caudate_abs <- phyloseq::subset_samples(absolute, Clade.Order == "Caudate")
Three_Genara_abs <- phyloseq::subset_samples(Caudate_abs, Host.genus %in% c("Desmognathus", "Plethodon", "Eurycea"))
Three_Genara_abs_BlueRidge <- phyloseq::subset_samples(Three_Genara_abs, Ecoregion.III == "Blue Ridge")
Desmog_Blue_Ins_16_abs <- phyloseq::subset_samples(Three_Genara_abs_BlueRidge, Host.genus == "Desmognathus")

results_DESeq2 <- perform_and_visualize_DA(
  obj = Desmog_Blue_Ins_16_abs,
  method = "DESeq2",
  group_var = "Host.taxon",
  contrast = c("Desmognathus monticola", "Desmognathus imitator"),
  output_csv_path = "DA_DESeq2.csv",
  target_glom = "Genus",
  significance_level = 0.05
)


head(results_DESeq2$results)
results_DESeq2$obj_significant
results_DESeq2$plot
results_DESeq2$bar_plot
```



```{r}
# Relative abundance
data("physeq_16SOTU", package = "DspikeIn")
physeq_16SOTU <- tidy_phyloseq_tse(physeq_16SOTU)

relative <- phyloseq::subset_taxa(physeq_16SOTU, Genus != "Tetragenococcus")
Caudate_rel <- phyloseq::subset_samples(relative, Clade.Order == "Caudate")
Three_Genara_rel <- phyloseq::subset_samples(Caudate_rel, Host.genus %in% c("Desmognathus", "Plethodon", "Eurycea"))
Three_Genara_rel_BlueRidge <- phyloseq::subset_samples(Three_Genara_rel, Ecoregion.III == "Blue Ridge")
Desmog_Blue_Ins_16_rel <- phyloseq::subset_samples(Three_Genara_rel_BlueRidge, Host.genus == "Desmognathus")

results_DESeq2_rel <- perform_and_visualize_DA(
  obj = Desmog_Blue_Ins_16_rel,
  method = "DESeq2",
  group_var = "Host.taxon",
  contrast = c("Desmognathus monticola", "Desmognathus imitator"),
  output_csv_path = "DA_DESeq2.csv",
  target_glom = "Genus",
  significance_level = 0.05
)

# head(results_DESeq2_rel$results) #  sig taxa
# results_DESeq2_rel$plot
# results_DESeq2_rel$bar_plot
# results_DESeq2_rel$obj_significant
```

# Multilayer Pairwise DA 

```{r}
#  for all levels

# Ensure Host.taxon is a factor
# sample_data(Desmog_Blue_Ins_16_rel)$Host.taxon <- as.factor(sample_data(Desmog_Blue_Ins_16_rel)$Host.taxon)
# # Get all unique levels
# host_levels <- levels(sample_data(Desmog_Blue_Ins_16_rel)$Host.taxon)
# # Build pairwise contrasts
# contrast_named <- list(
#   Host.taxon = combn(levels(sample_data(Desmog_Blue_Ins_16_rel)$Host.taxon), 2, simplify = FALSE)
# )

# select some levels
contrast_named <- list(
  Host.taxon = list(
    c("Desmognathus adatsihi", "Desmognathus aeneus"),
    c("Desmognathus adatsihi", "Desmognathus conanti"),
    c("Desmognathus adatsihi", "Desmognathus imitator")
  )
)

multi_lev_results <- perform_and_visualize_DA(
  obj = Desmog_Blue_Ins_16_rel,
  method = "DESeq2",
  contrast = contrast_named,
  target_glom = "Genus",
  global_fdr = TRUE,
  output_csv_path = "DA_multi.csv"
)

multi_lev_results$Host.taxon_Desmognathus.adatsihi_vs_Desmognathus.aeneus$bar_plot
multi_lev_results$Host.taxon_Desmognathus.adatsihi_vs_Desmognathus.conanti$bar_plot
multi_lev_results$Host.taxon_Desmognathus.adatsihi_vs_Desmognathus.imitator$bar_plot
```



#                     Visualization


```{r}
# ===========================================================
#  Visualization of community composition
# ===========================================================


Rel <- phyloseq::subset_taxa(physeq_16SOTU, Genus != "Tetragenococcus")
Prok_OTU_spiked <- phyloseq::subset_samples(Rel, spiked.volume %in% c("2", "1"))
Prok_OTU_spiked <- phyloseq::subset_samples(Prok_OTU_spiked, sample.or.blank != "blank")
Prok_OTU_sal <- phyloseq::subset_samples(Prok_OTU_spiked, Animal.type == "Salamander")

Prok_OTU_sal <- tidy_phyloseq_tse(Prok_OTU_sal)

taxa_barplot(
  Prok_OTU_sal,
  target_glom = "Genus",
  custom_tax_names = NULL,
  normalize = TRUE,
  treatment_variable = "Habitat",
  abundance_type = "relative",
  x_angle = 25,
  fill_variable = "Family",
  facet_variable = "Diet",
  top_n_taxa = 20,
  palette = DspikeIn::color_palette$MG,
  legend_size = 11,
  legend_columns = 1,
  x_scale = "free",
  xlab = NULL
)
```




#                 Microbial dynamics & NetWork comparision


```{r}
# ===========================================================
# 1. Initialization and loading NetWorks for Comparision
# ===========================================================

# library(SpiecEasi)
# library(ggnet)
# library(igraph)
library(DspikeIn)
library(tidyr)
library(dplyr)
library(ggpubr)
library(igraph)

# To create a microbial co-occurrence network, you can refer to the SpiecEasi package available at:
# SpiecEasi GitHub Repository  https://github.com/zdk123/SpiecEasi

# herp.Bas.rel.f is a merged phyloseq object for both bacterial and fungal domains
# spiec.easi(herp.Bas.rel.f, method='mb', lambda.min.ratio=1e-3, nlambda=250,pulsar.select=TRUE )

Complete <- load_graphml("Complete.graphml")
NoBasid <- load_graphml("NoBasid.graphml")
NoHubs <- load_graphml("NoHubs.graphml")

# result <- weight_Network(graph_path = "Complete.graphml")
# result

result_kk <- degree_network(
  graph_path = load_graphml("Complete.graphml"),
  save_metrics = TRUE,
  layout_type = "stress"
)
# print(result_kk$plot)
```





#            Network Topological Metrics


```{r}
# ===========================================================
# 2. Metrics Calculation
# ===========================================================

result_Complete <- node_level_metrics(Complete)
result_NoHubs <- node_level_metrics(NoHubs)
result_NoBasid <- node_level_metrics(NoBasid)

Complete_metrics <- result_Complete$metrics
Nohub_metrics <- result_NoHubs$metrics
Nobasid_metrics <- result_NoBasid$metrics

Complete_metrics <- data.frame(lapply(Complete_metrics, as.character), stringsAsFactors = FALSE)
Nohub_metrics <- data.frame(lapply(Nohub_metrics, as.character), stringsAsFactors = FALSE)
Nobasid_metrics <- data.frame(lapply(Nobasid_metrics, as.character), stringsAsFactors = FALSE)


print(igraph::vcount(Complete)) # Number of nodes
print(igraph::ecount(Complete)) # Number of edges

print(igraph::vcount(NoBasid))
print(igraph::ecount(NoBasid))

print(igraph::vcount(NoHubs))
print(igraph::ecount(NoHubs))


metrics_scaled <- bind_rows(
  Complete_metrics %>% mutate(Network = "Complete"),
  Nohub_metrics %>% mutate(Network = "NoHubs"),
  Nobasid_metrics %>% mutate(Network = "NoBasid")
) %>%
  dplyr::mutate(dplyr::across(where(is.numeric), scale))

metrics_long_scaled <- metrics_scaled %>%
  tidyr::pivot_longer(cols = -c(Node, Network), names_to = "Metric", values_to = "Value")
```



bind the metrics to plot them


```{r}
# ===========================================================
# 3. Visualization
# ===========================================================

# Remove missing values
metrics_long_scaled <- na.omit(metrics_long_scaled)

# We visualize only six metrics
selected_metrics <- c(
  "Degree", "Closeness", "Betweenness",
  "EigenvectorCentrality", "PageRank", "Transitivity"
)

metrics_long_filtered <- metrics_long_scaled %>%
  filter(Metric %in% selected_metrics) %>%
  mutate(
    Value = as.numeric(as.character(Value)),
    Network = recode(Network,
      "Complete" = "Complete Network",
      "NoHubs" = "Network & Module Hubs Removed",
      "NoBasid" = "Basidiobolus Subnetwork Removed"
    )
  ) %>%
  na.omit() # Remove any NA if any

metrics_long_filtered$Network <- factor(metrics_long_filtered$Network,
  levels = c(
    "Complete Network",
    "Network & Module Hubs Removed",
    "Basidiobolus Subnetwork Removed"
  )
)

# DspikeIn::color_palette$light_MG
network_colors <- c(
  "Complete Network" = "#F1E0C5",
  "Network & Module Hubs Removed" = "#D2A5A1",
  "Basidiobolus Subnetwork Removed" = "#B2C3A8"
)

# statistical comparisons a vs b
comparisons <- list(
  c("Complete Network", "Network & Module Hubs Removed"),
  c("Complete Network", "Basidiobolus Subnetwork Removed"),
  c("Network & Module Hubs Removed", "Basidiobolus Subnetwork Removed")
)

networks_in_data <- unique(metrics_long_filtered$Network)
comparisons <- comparisons[sapply(comparisons, function(pair) all(pair %in% networks_in_data))]

ggplot(metrics_long_filtered, aes(x = Network, y = Value, fill = Network)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(aes(color = Network),
    position = position_jitter(0.2), alpha = 0.2, size = 1.5
  ) +
  scale_fill_manual(values = network_colors) +
  scale_color_manual(values = network_colors) +
  facet_wrap(~Metric, scales = "free_y", labeller = label_wrap_gen(width = 20)) +
  ggpubr::stat_compare_means(method = "wilcox.test", label = "p.signif", comparisons = comparisons) +
  theme_minimal() +
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.x = element_text(size = 10, angle = 10, hjust = 0.8),
    strip.text = element_text(size = 12, margin = margin(t = 15, b = 5)),
    legend.position = "top",
    legend.text = element_text(size = 13),
    legend.title = element_text(size = 13, face = "bold"),
    plot.title = element_text(size = 13, face = "bold")
  ) +
  labs(title = "Selected Node Metrics Across Networks", fill = "Network Type", color = "Network Type")
```


#          To find first and second neighbors your target node  


```{r}
Complete <- load_graphml("Complete.graphml")
result2 <- extract_neighbors(
  graph = Complete,
  target_node = "OTU69:Basidiobolus_sp", mode = "all"
)
head(result2$summary)
```



#                 Compute Node-Level Metrics

 | **Metric**                     | **Description** |
 |--------------------------------|-----------------------------------------------------------|
 | 'Node'                         | Node name (character format) |
 | 'Degree'                       | Number of edges connected to the node |
 | 'Strength'                     | Sum of edge weights connected to the node |
 | 'Closeness'                    | Closeness centrality (normalized, based on shortest paths) |
 | 'Betweenness'                  | Betweenness centrality (normalized, measures control over network flow) |
 | 'EigenvectorCentrality'        | Eigenvector centrality (importance based on connections to influential nodes) |
 | 'PageRank'                     | PageRank score (importance based on incoming links) |
 | 'Transitivity'                 | Local clustering coefficient (tendency of a node to form triangles) |
 | 'Coreness'                     | Node's coreness (from k-core decomposition) |
 | 'Constraint'                   | Burt's constraint (measures structural holes in a node's ego network) |
 | 'EffectiveSize'                | Inverse of constraint (larger values = more non-redundant connections) |
 | 'Redundancy'                   | Sum of constraint values of a node's alters |
 | 'Community'                    | Community assignment from Louvain clustering |
 | 'Efficiency'                   | Global efficiency (average inverse shortest path length) |
 | 'Local_Efficiency'             | Local efficiency (subgraph efficiency for a node's neighbors) |
 | 'Within_Module_Connectivity'   | Proportion of neighbors in the same community |
 | 'Among_Module_Connectivity'    | Proportion of neighbors in different communities |
 
 
```{r}
# Load required libraries
library(igraph)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggrepel)

# Compute Node-Level Metrics
completeMetrics <- node_level_metrics(Complete)
NoHubsMetrics <- node_level_metrics(NoHubs)
NoBasidMetrics <- node_level_metrics(NoBasid)

completeMetrics$facet_plot

# Ensure each dataset has a "Network" column before combining
completeMetrics$metrics <- completeMetrics$metrics %>% mutate(Network = "Complete Network")
NoHubsMetrics$metrics <- NoHubsMetrics$metrics %>% mutate(Network = "Network & Module Hubs Removed")
NoBasidMetrics$metrics <- NoBasidMetrics$metrics %>% mutate(Network = "Basidiobolus Subnetwork Removed")

# Combine All Data
combined_data <- bind_rows(
  completeMetrics$metrics,
  NoHubsMetrics$metrics,
  NoBasidMetrics$metrics
)


# Add Node Identifier if missing
if (!"Node" %in% colnames(combined_data)) {
  combined_data <- combined_data %>% mutate(Node = rownames(.))
}

# Convert `Network` into Factor
combined_data$Network <- factor(combined_data$Network, levels = c(
  "Complete Network",
  "Network & Module Hubs Removed",
  "Basidiobolus Subnetwork Removed"
))

# Convert Data to Long Format
metrics_long <- combined_data %>%
  pivot_longer(
    cols = c("Redundancy", "Efficiency", "Betweenness"),
    names_to = "Metric", values_to = "Value"
  )

# Define Custom Colors and Shapes
network_colors <- c(
  "Complete Network" = "#F1E0C5",
  "Network & Module Hubs Removed" = "#D2A5A1",
  "Basidiobolus Subnetwork Removed" = "#B2C3A8"
)

network_shapes <- c(
  "Complete Network" = 21,
  "Network & Module Hubs Removed" = 22,
  "Basidiobolus Subnetwork Removed" = 23
)

# Determine Top 30% of Nodes to Label/Optional
metrics_long <- metrics_long %>%
  group_by(Network, Metric) %>%
  mutate(Label = ifelse(rank(-Value, ties.method = "random") / n() <= 0.3, Node, NA))

# ?quadrant_plot() can creat plot for indivisual network
# plot <- quadrant_plot(metrics, x_metric = "Degree", y_metric = "Efficiency")

# Create comparision Plots
create_metric_plot <- function(metric_name, data, title) {
  data_filtered <- data %>% filter(Metric == metric_name)
  median_degree <- median(data_filtered$Degree, na.rm = TRUE)
  median_value <- median(data_filtered$Value, na.rm = TRUE)

  ggplot(data_filtered, aes(x = Degree, y = Value, fill = Network)) +
    geom_point(aes(shape = Network), size = 3, stroke = 1, color = "black") +
    geom_text_repel(aes(label = Label), size = 3, max.overlaps = 50) +
    scale_fill_manual(values = network_colors) +
    scale_shape_manual(values = network_shapes) +
    geom_vline(xintercept = median_degree, linetype = "dashed", color = "black", size = 1) +
    geom_hline(yintercept = median_value, linetype = "dashed", color = "black", size = 1) +
    labs(
      title = title,
      x = "Degree",
      y = metric_name,
      fill = "Network",
      shape = "Network"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(
        hjust = 0.5, size = 16, face = "bold",
        margin = margin(t = 10, b = 20) # Moves the title downward
      ),
      axis.title = element_text(size = 14, face = "bold"),
      legend.position = "top",
      legend.title = element_text(size = 14, face = "bold"),
      legend.text = element_text(size = 12)
    )
}


# Generate Plots
plot_redundancy <- create_metric_plot("Redundancy", metrics_long, "Redundancy vs. Degree Across Networks")
plot_efficiency <- create_metric_plot("Efficiency", metrics_long, "Efficiency vs. Degree Across Networks")
plot_betweenness <- create_metric_plot("Betweenness", metrics_long, "Betweenness vs. Degree Across Networks")

# Save Plots
# ggsave("plot_redundancy_20percent.png", plot_redundancy, width = 8, height = 6)
# ggsave("plot_efficiency_20percent.png", plot_efficiency, width = 8, height = 6)
# ggsave("plot_betweenness_20percent.png", plot_betweenness, width = 8, height = 6)

# Print Plots
# print(plot_redundancy)
# print(plot_efficiency)
# print(plot_betweenness)
```



#            Compute degree metrics and visualize the network


```{r}
# Compute degree metrics and visualize the network
# Options: `"stress"` (default), `"graphopt"`, `"fr"`

result <- degree_network(graph_path = Complete, save_metrics = TRUE)
print(result$metrics)
print(result$plot)


# Compute network weights for different graph structures
NH <- weight_Network(graph_path = "NoHubs.graphml")
NB <- weight_Network(graph_path = "NoBasid.graphml")
C <- weight_Network(graph_path = "Complete.graphml")

# Extract metrics from the computed network weights
CompleteM <- C$metrics
NoHubsM <- NH$metrics
NoBasidM <- NB$metrics

# Combine metrics into a single dataframe for comparison
df <- bind_rows(
  CompleteM %>% mutate(Group = "CompleteM"),
  NoHubsM %>% mutate(Group = "NoHubsM"),
  NoBasidM %>% mutate(Group = "NoBasidM")
) %>%
  pivot_longer(cols = -Group, names_to = "Metric", values_to = "Value")

# Aggregate the total values by metric and group
df_bar <- df %>%
  group_by(Metric, Group) %>%
  summarise(Total_Value = sum(Value), .groups = "drop")

# Plot the metrics comparison
ggplot(df_bar, aes(x = Metric, y = log1p(Total_Value), fill = Group)) +
  geom_bar(stat = "identity", position = "dodge", alpha = 0.8) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_manual(values = c("#F1E0C5", "#D2A5A1", "#B2C3A8")) +
  labs(
    title = "Total Network Metrics Comparison",
    x = "Metric",
    y = "Total Value (log-scaled)",
    fill = "Group"
  )
```


```{r}
sessionInfo()
```


```{r cleanup, include = FALSE}
# =====================================================================
# FINAL CLEANUP: Remove files created during vignette execution
# =====================================================================

# **Note:** To comply with CRAN policies, this vignette removes any output files
# (e.g., plots, summary tables) created during execution.
# You may modify paths locally if you'd like to keep the outputs.

# List of all files possibly written by example functions
files_to_delete <- c(
  "Global_Network_Metrics.csv", "absolute_counts.csv",
  "abundance_analysis_filtered_phyloseq.rds", "abundance_analysis_filtered_tse.rds",
  "abundance_analysis_high_abundance_phyloseq.rds", "abundance_analysis_high_abundance_tse.rds",
  "abundance_analysis_low_abundance_phyloseq.rds", "abundance_analysis_low_abundance_tse.rds",
  "adjusted_prevalence.rds", "core_microbiome.csv", "core_microbiome.rds", "comp_spikein_patristic_distances.csv", "comp_spikein_branch_lengths.pdf", "comp_spikein_patristic_distances_hist.pdf",
  "merged_data.csv", "merged_data.docx", "comp_spikein.nwk", "comp_spikein_tree.png", "comp_spikein_patristic_distances_hist.pfd",
  "merged_physeq_max_processed.rds", "merged_physeq_sum.rds",
  "merged_physeq_sum_processed.rds", "spikeIn_factors_output",
  "plot_Spike.percentage_Kruskal.pdf", "plot_Spike.percentage_Kruskal.png",
  "plot_Spike.reads_Kruskal.pdf", "plot_Spike.reads_Kruskal.png",
  "plot_Total.reads_Kruskal.pdf", "plot_Total.reads_Kruskal.png",
  "proportion_adjusted_physeq.rds", "proportion_adjusted_tse.rds",
  "randomsubsample_Trimmed_evenDepth.rds", "summary.csv", "summary.docx", "DspikeIn-manual.tex", "output.csv", "output.docx",
  "tree_with_bootstrap_and_cophenetic.png"
)

# Remove files safely
invisible(file.remove(files_to_delete[file.exists(files_to_delete)]))

# Remove known folders (if created by functions)
folders_to_delete <- c("spikeIn_factors_output")
for (folder in folders_to_delete) {
  if (dir.exists(folder)) {
    unlink(folder, recursive = TRUE, force = TRUE)
  }
}
```


#         DspikeIn volume protocol

Spike-in volume Protocol; 

The species Tetragenococcus halophilus (bacterial spike; ATCC33315) and Dekkera bruxellensis (fungal spike; WLP4642-White Labs) were selected as taxa to spike into gut microbiome samples as they were not found in an extensive collection of wildlife skin (GenBank BioProjects: PRJNA1114724, PRJNA 1114659) or gut microbiome samples. Stock cell suspensions of both microbes were grown in either static tryptic soy broth (T. halophilus) or potato dextrose broth (D. bruxellensis) for 72 hours then serially diluted and optical density (OD600) determined on a ClarioStar plate reader. Cell suspensions with an optical density of 1.0, 0.1, 0.01, 0.001 were DNA extracted using the Qiagen DNeasy Powersoil Pro Kit. These DNA isolations were used as standards to determine the proper spike in volume of cells to represent 0.1-10% of a sample (Rao et al., 2021b) Fecal pellets (3.1 ± 1.6 mg; range = 1 – 5.1 mg) from an ongoing live animal study using wood frogs (Lithobates sylvaticus) were used to standardize the input material for the development of this protocol. A total of (n=9) samples were used to validate the spike in protocol. Each fecal sample was homogenized in 1mL of sterile molecular grade water then 250uL of fecal slurry was DNA extracted as above with and without spiked cells. 
Two approaches were used to evaluate the target spike-in of 0.1-10%, the range of effective spike-in percentage described in (Rao et al., 2021b), including 1) an expected increase of qPCR cycle threshold (Ct) value that is proportional to the amount of spiked cells and 2) the expected increase in copy number of T. halophilus and D. bruxellensis in spiked vs. unspiked samples. A standard curve was generated using a synthetic fragment of DNA for the 16S-V4 rRNA and ITS1 rDNA regions of T. halophilus and D. bruxellensis, respectively. The standard curve was used to convert Ct values into log copy number for statistical analyses (detailed approach in[2, 3]) using the formula y = -0.2426x + 10.584 for T. halophilus and y = -0.3071x + 10.349 for D. bruxellensis, where x is the average Ct for each unknown sample.
Quantitative PCR (qPCR) was used to compare known copy numbers from synthetic DNA sequences of T. halophilus and D. bruxellensis to DNA extractions of T. halophilus and D. bruxellensis independently, and wood frog fecal samples with and without spiked cells. SYBR qPCR assays were run at 20ul total volume including 10ul 2X Quantabio PerfeCTa SYBR Green Fastmix, 1ul of 10uM forward and reverse primers, 1ul of ArcticEnzymes dsDNAse master mix clean up kit, and either 1ul of DNA for D. bruxellensis or 3ul for T. halophilus. Different volumes of DNA were chosen for amplification of bacteria and fungi due to previous optimization of library preparation and sequencing steps [3]. The 515F [4] and 806R [5] primers were chosen to amplify bacteria and ITS1FI2 [6] and ITS2 for fungi, as these are the same primers used during amplicon library preparation and sequencing. Cycling conditions on an Agilent AriaMX consisted of 95 C for 3 mins followed by 40 cycles of 95 C for 15 sec, 60 C for 30 sec and 72 C for 30 sec. Following amplification, a melt curve was generated under the following conditions including 95 C for 30 sec, and a melt from 60 C to 90 C increasing in resolution of 0.5 C in increments of a 5 sec soak time. To validate the spike in protocol we selected two sets of fecal samples including 360 samples from a diverse species pool of frogs, lizards, salamanders and snakes and a more targeted approach of 122 fecal samples from three genera of salamanders from the Plethodontidae. (Supplemental Table #). FFecal samples were not weighed in the field, rather, a complete fecal pellet was diluted in an equal volume of sterile water and standardized volume of fecal slurry (250µL) extracted for independent samples.. A volume of 1ul T. halophilus (1847 copies) and 1ul D. bruxellensis (733 copies) were spiked into each fecal sample then DNA was extracted as above, libraries constructed, and amplicon sequenced on an Illumina MiSeq as in [7] .

<https://github.com/mghotbi/DspikeIn>
